---
title: "Sensitivity_depth_analysis"
author: "Torbjörn Säterberg"
date: today  
format: html
embed-resources: true
self-contained: true
editor: visual
code-fold: true
error: FALSE
warning: FALSE
message: FALSE
toc: true
toc-depth: 3
lang: en  
---

# 1. Introduction

This script is used as a sensitivity analysis of the effect of depth on mortality in the Säterberg et al. 2026 perch model paper.

As there are indications that large perch tend to in habituate shallow waters to a larger extent than what smaller perch do ([Bergström et al. 2016](https://www.sciencedirect.com/science/article/pii/S0272771416304899%5D)) we subset the data based on water depth at which gill-nets are set and redo the analysis using various data subsets. The analysis is conducted in order to test whether the age-mortality pattern observed among coastal perch populations persist when only shallow sites are used as data source. If the same pattern, that is, an increasing mortality with age, is observed for shallow as well as deep areas it indicates that the population is relatively well sampled.

## 1.1 Required R-packages {#R-packages}

```{r}
#| label: load-packages
require("RPostgres", quietly = TRUE) # postgres functionality in R
library(DBI) # For database connection
library(odbc) # Connection between R and relation database management systems
library(ggforce, quietly=TRUE)  # added functionality to ggplot
library(tidybayes) # working with posterior draws for bayesian models
library(flextable) # plotting nice tables
library(brms) # multilevel models in a bayesian context(similar to lmer)
library(bayesplot) # plotting bayesian models
library(future) # parallell computing 
library(purrr) # working with functions in pipe
library(modelr) # modeling functions with pipe
library(tidyverse) # tidyverse functionality for working with data
library(cmdstanr) # fitting stan models through R
library(patchwork) # combining multiple ggplots
```

## 1.2 Connect to KUL-database {#dbconnection}

```{r}
# Connect to Aqua-datalagret
AQ_con <- dbConnect(odbc(), 
                         Driver = "SQL Server", 
                         Server = "aquadw_prod.db.slu.se",
                         USER = Sys.info()["user"], 
                         Database = "AquaDW_prod", 
                         Trusted_Connection = "TRUE", 
                         encoding = "UTF-8")
```

## 1.3 Source utility functions

```{r}
files <- list.files("../R", pattern = "\\.R$", full.names = TRUE)
invisible(lapply(files, function(f) {
  source(f, echo = FALSE, verbose = FALSE)
}))
```

# 2. Load data

Here we load data from the KUL data base. Data is extracted from three different views(tables) that are needed in the assessment model. For perch we are only interested in gill-net nights that are undisturbed (störing=NEJ), approved (Godkänd=JA) and are sampled at 0-10 m depth. The tables we use are the following:

-   KUL_OBSERVATION_NEW
    -   This table includes information on effort.
-   KUL_LANGDGRUPP
    -   This table includes the number of individuals caught per length class
-   KUL_INDIVID_NEW
    -   This table includes data on disected individuals(length and age measurements of females)
-   KUL_STICKPROVSUTFALL
    -   This table includes information on number of females/males in the sample. This information is needed in order to find the number of females in the stocks.

We load data using a direct read-only database connection to SLU datalagret and are only interested in K064 data series(Nordiska kustöversiktsnät). We load data from the following test fishing areas:

-   Torhamn
-   Holmön
-   Forsmark
-   Långvindsfjärden
-   Norrbyn
-   Kinnbäcksfjärden
-   Gaviksfjärden
-   Lagnö
-   Råneå
-   Asköfjärden
-   Kvädöfjärden
-   Finbo (Åland)
-   Kumlinge (Åland)
-   Askrikefjärden (Vaxholm)

## 2.1 Load effort data {#effordata}

Here we load and summarize effort data from the KUL table "KUL_OBSERVATION_NEW". For perch we are only interested in gill-net nights that are undisturbed (störing=NEJ), approved (Godkänd=JA) and are sampled at 0-10 m depth.

```{sql, connection="AQ_con", output.var = "effort"}
-- CALCULATE EFFORT
SELECT Lokal, Fångstområde, År, FISKE_ID, OBS_ID, DJUPSTRATUM AS Djupstratum
FROM AquaDW_prod.mart.KUL_OBSERVATION_NEW
WHERE 
Redskap_kod = 'K064' 
AND 
Störning='NEJ'
AND 
Godkänd ='JA'
AND 
Lokal_ID IN(129, 49, 31, 65, 79, 84, 82, 66, 106, 94, 54, 28, 57, 203) -- Torhamn, Holmön, Forsmark, Långvindsfjärden, Norrbyn, Kinnbäcksfjärden, Gaviksfjärden, Lagnö, Råneå, Asköfjärden, Kvädöfjärden, Finbo(Åland), Kumlinge(Åland), Askrikefjärden(Vaxholm)
AND 
Fiske IN ('Nordiska nät augusti','Kontrollprogram Nordiska nät augusti')
AND 
Fångstområde <> 'Långvindsfjärden reservat'
AND 
STRATIFIERING NOT IN ('Djupstratum(0-10-20-30): 10-20 m', -- remove data from deepest strata
'Djupstratum: 10-20 m', 
'Djupstratum: >20 m'); 
```

## 2.2 Load Catch data {#catchdata}

Here we load catch data for length groups. For perch we are only interested in gill-nets that are undisturbed (störing=NEJ), approved (Godkänd=JA) and are sampled at 0-10 m.

```{sql, connection="AQ_con", output.var = "Kul_data"}
SELECT *
FROM AquaDW_prod.mart.KUL_LANGDGRUPP klmt --
WHERE 
Redskap_kod = 'K064' 
AND
Art = 'Abborre' 
AND
Godkänd='JA'
AND 
Störning='NEJ'
AND 
Lokal_ID IN(129, 49, 31, 65, 79, 84, 82, 66, 106, 94, 54, 28, 57, 203) -- Torhamn, Holmön, Forsmark, Långvindsfjärden, Norrbyn, Kinnbäcksfjärden, Gaviksfjärden, Lagnö, Råneå, Asköfjärden, Kvädöfjärden, Finbo(Åland), Kumlinge(Åland), Askrikefjärden(Vaxholm)
AND 
Fiske IN ('Nordiska nät augusti','Kontrollprogram Nordiska nät augusti')
AND 
Fångstområde <> 'Långvindsfjärden reservat'
AND
Stratifiering NOT IN ('Djupstratum(0-10-20-30): 10-20 m',
'Djupstratum: 10-20 m', 
'Djupstratum: >20 m')
```

## 2.3 Load individual data {#inddata}

Here we load individual data. Length and age of individual fish.

```{sql, connection="AQ_con", output.var = "Kul_ind"}
SELECT * 
FROM
AquaDW_prod.mart.KUL_INDIVID_NEW
WHERE
Art = 'Abborre'
AND 
FISKE_ID IN
(SELECT DISTINCT FISKE_ID
FROM AquaDW_prod.mart.KUL_LANGDGRUPP klmt --
WHERE 
Redskap_kod = 'K064' 
AND
Art = 'Abborre' 
AND
Godkänd='JA'
AND 
Störning='NEJ'
AND 
Lokal_ID IN(129, 49, 31, 65, 79, 84, 82, 66, 106, 94, 28, 57, 203) -- Torhamn, Holmön, Forsmark, Långvindsfjärden, Norrbyn, Kinnbäcksfjärden, Gaviksfjärden, Lagnö, Råneå, Asköfjärden, Finbo(Åland), Kumlinge(Åland), Askrikefjärden(Vaxholm)
AND 
Fiske IN ('Nordiska nät augusti','Kontrollprogram Nordiska nät augusti')
AND 
Fångstområde <> 'Långvindsfjärden reservat'
AND
Stratifiering NOT IN ('Djupstratum(0-10-20-30): 10-20 m',
'Djupstratum: 10-20 m', 
'Djupstratum: >20 m')
)
OR -- In Kvädöfjärden individuals are sampled using nätlänkar some years and nordiska nät other years.
FISKE_ID IN 
(SELECT DISTINCT FISKE_ID
FROM
AquaDW_prod.mart.KUL_INDIVIDPROV
WHERE
Art = 'Abborre'
AND 
Lokal_ID IN(54) 
AND 
Fiske IN('Nordiska nät augusti', 'Nätlänk augusti'));
```

## 2.4 Load sex data {#sexdata}

Here we load information on proportion of males and females in subsamples of the data.

```{sql, connection="AQ_con", output.var = "Kul_sex"}
SELECT Lokal, Fiskeår, Längdgrupp, ANTAL_HANAR, ANTAL_HONOR, ANTAL
FROM AquaDW_prod.mart.KUL_STICKPROVSUTFALL ks
WHERE
Lokal_ID IN(129, 49, 31, 65, 79, 84, 82, 66, 106, 94, 54, 28, 57, 203) -- Torhamn, Holmön, Forsmark, Långvindsfjärden, Norrbyn, Kinnbäcksfjärden, Gaviksfjärden, Lagnö, Råneå, Asköfjärden, Kvädöfjärden, Finbo(Åland), Kumlinge(Åland), Askrikefjärden(Vaxholm)
AND 
Fiske IN ('Nordiska nät augusti','Kontrollprogram Nordiska nät augusti');
```

# 3 Transform data {#transformdata}

Here we transform and analyse data. Age-length keys are derived, proportion of females in the samples are calculated and data is subsetted into age-classes.

## 3.1 Remove NAs from individual and catch data sets {#filterdata}

```{r}
Kul_data <- Kul_data %>% 
            filter(!is.na(Längdgrupp),!is.na(Antal)) %>% # remove cases with no length info
            uncount(Antal, .remove = FALSE) # expand df such that each row is one individual

Kul_ind <- Kul_ind %>% 
           filter(!is.na(Slutlängd_cm),!is.na(Ålder)) # we cant use NAs
```

## 3.2 Estimate age based indices using ALK

In this section we:

1.  [Derive age-length keys](#ALKs)([ALK](#ALK-function)) using [age-read individuals](#inddata)
2.  [Estimate age of individual fish](#estfishage) using [ALKs](#ALKs) and the [full data set with fish lengths](#catchdata).
3.  [Summarize catches per age class](#fishperlengthclass)
4.  [Restructure the summarized catches (3) per age into cohorts](#cohort).

### 3.2.1 ALK function {#ALK-function}

This function is used for deriving age-length keys for data subsets. It takes as input the observed lengths in [the total catch](#catchdata) (*lengthD*) and [age and length data](#inddata) from a subsample of the total catch to derive an age-length key (i.e. the probability of age given length). This is sometimes called a forward age-length key [(Ailloud & Hoenig 2019 ICES J Sci)](https://academic.oup.com/icesjms/article/76/6/1515/5480389).

```{r}
applyALK <- function(lengthD,indD){

# Find number of individuals per length and age in individual data
LADat <- xtabs(~Slutlängd_cm + Ålder, data=indD) # number of observations per age- and length-class 
ind_lidx <- sort(unique(indD$Slutlängd_cm)) # length groups in individual data
L_lidx <- sort(unique(lengthD$Längdgrupp)) # Length groups in length data

# initialize Length/age group matrix which will be fille later
l_agr_mat <- matrix(NA,length(L_lidx),dim(LADat)[2])
rownames(l_agr_mat) <- L_lidx 
colnames(l_agr_mat) <- colnames(LADat)
l_agr_mat[L_lidx %in% ind_lidx,] <- LADat[rownames(l_agr_mat)[L_lidx %in% ind_lidx],]
rowidNA <- as.numeric(rownames(l_agr_mat)[apply(l_agr_mat, 1, anyNA)]) # columns to fill with data

# Fill in values for length groups with no individual data (borrow age data from lower or higher length-classes)
for(i in rowidNA){
  if(i<min(ind_lidx)){
  l_agr_mat[rownames(l_agr_mat)==as.character(i),] <- LADat[as.character(min(ind_lidx)),] # take the smallest length group from ind data if Length observations are smaller than individual data
  } else if (i>max(ind_lidx)) {
  l_agr_mat[rownames(l_agr_mat)==as.character(i),] <- LADat[as.character(max(ind_lidx)),] # take age dist from the largest length group from ind data if Length observations are larger than individual data 
  }else{    
  # take the sum of age distributions for closest length groups if length group is not part of individual data 
  l_agr_mat[rownames(l_agr_mat)==as.character(i),] <- colSums(LADat[c(as.character(min(ind_lidx[ind_lidx>i])), 
                                                          as.character(max(ind_lidx[ind_lidx<i]))),])
  }
  }

# convert from number of ind per age- and length class to proportion of age-classes per length-class in sample.
ALK <- prop.table(l_agr_mat,margin=1)
return(ALK)
}
```

### 3.2.2 Derive ALKs for each subset of data (FISKE_ID) {#ALKs}

Age-length keys are here derived for each subset of data(FISKE_ID) using data of [indivual age-read and length measured fish](#inddata) in combination with the [full data set](#catchdata). The full data set is only used to give information on the number of length-classes needed in the age-length keys.

```{r}

# Apply function to grouped dataframes
ALK_all <- Kul_data %>%
  group_by(Lokal, År) %>% # Group length data from full sample using FISKE_ID
  nest() %>%  
  inner_join( # Append individual data to full length sample data set group wise
    Kul_ind %>%
      group_by(Lokal, År) %>%
      nest(),
    by = c("Lokal","År")
  ) %>%
  mutate(
    ALK = purrr::map2(data.x, data.y, applyALK) # derive ALK for each subset of data
  ) %>%
    ungroup() %>%
  select(-data.x, -data.y)
```

### 3.2.3 Estimate age of individual fish {#estfishage}

Here we use ALK and the full data set, subsetted by FISKE_ID, to estimate age of individual fish in the full length data set.

```{r}
# For reproducibility set seed
set.seed(123)

# sample individual age across all individuals (assuming only females in the sample)
ind_age <- Kul_data %>%
  group_by(Lokal,Djupstratum, År) %>%
  nest() %>%
  inner_join(
    ALK_all %>%
      group_by(Lokal,År),
    by = c("Lokal","År")
  ) %>%
  mutate(
    ind_age = map2(data, ALK, function(x,y) {FSA::alkIndivAge(y,~ Längdgrupp,data=x)$age})
  ) %>%
  unnest(c(ind_age, data)) %>% 
  ungroup() %>%
  select(Lokal, Djupstratum, År, OBS_ID, Längdgrupp, ind_age)

```

## 3.3 Assign sex to individuals

1.  First we calculate the [proportion of females](#female_prop_area) per area from annual sub-samples.
2.  Second, we [randomly assign sex to individuals](#assignsex) based on the calculated proportion of females per area and length group.

### 3.3.1 Proportion females per area (i.e. aggregate data across years for each area) {#female_prop_area}

Here we aggregate data across years in order to calculate the proportion of females per length group and area.

```{r}

# female proportions per area and year
female_prop_Y <- Kul_sex %>% 
  mutate(females = case_when(ANTAL_HANAR==ANTAL ~ 0, # Number of males equal to total -> 0 females
                             !is.na(ANTAL_HONOR) ~ ANTAL_HONOR, # Number of females reported
                             .default = NA),
         total = case_when(!is.na(ANTAL) ~ ANTAL, # Total sample reported
                           (is.na(ANTAL)) & 
                           (!is.na(ANTAL_HANAR) & !is.na(ANTAL_HONOR)) 
                           ~ ANTAL_HANAR+ANTAL_HONOR,# Females and males reported -> prop can be calculated
                           .default=NA),
         female_prop=females/total # proportion females in size class from sample
                                  )

# femle proportions per area
female_prop_area <- 
  female_prop_Y %>% 
  filter(!if_any(c("females","total"),is.na)) %>% # remove cases where either females or the total number individuals in samples is na
  group_by(Lokal, Längdgrupp) %>%
  summarize(total = sum(total), 
            females = sum(females),
            prop_females = sum(females)/sum(total)) %>%
  ungroup()

```

### 3.3.2 Randomly assign sex to indivuals {#assignsex}

Here we randomly assign sex to individuals using a bernoulli model:

$$
X_{ijkl} \sim \text{Bern}(p_{jk})
$$

where $X_{ijk}$ is a binary random variable and ($p_{jk}$) is the probability that an individual perch in area *j* and age class *k* is a female. $p_{jk}$ is estimated ($\hat{p}_{jk}$) using [sex proportions](#female_prop_area) observed in samples of [sex](#sexdata).

```{r}
# for reproducibility set seed
set.seed(123)

# join sex probabilities to individual data
ind_sex <- ind_age %>%
  left_join(female_prop_area,by=c("Lokal","Längdgrupp"))

# randomly sample sex based on sex probabilities per length group
ind_sex$isfemale <- rbinom(n=nrow(ind_sex),size=1,p=ind_sex$prop_females)

# The previous line produces a few na:s for age groups that do not occur in the sex data. For these lines assign females to large individuals
ind_sex$isfemale[is.na(ind_sex$isfemale) & (ind_sex$Längdgrupp>30)] <- 1

# for small ones randomly sample sex based on 50-50 chance of male-female
ind_sex$isfemale[is.na(ind_sex$isfemale)] <- rbinom(n=sum(is.na(ind_sex$isfemale)),size=1,p=0.5) 
  
```

### 3.2.4 Summarize number of fish per age-class {#fishperageclass}

```{r}
age_per_obsid <- ind_sex %>%
  filter(!(År <= 2004 & Lokal=="Forsmark")) %>%  # remove odd years with data for forsmark
  group_by(Lokal, Djupstratum, År, OBS_ID, ind_age) %>% 
  summarize(N=sum(isfemale)) %>% # number of females per age and net
  ungroup() %>%
  complete(ind_age, nesting(OBS_ID, Djupstratum), fill=list(N=0)) %>%
  group_by(OBS_ID) %>% 
  mutate(Lokal=if_else(is.na(Lokal),unique(Lokal[!is.na(Lokal)])[1],Lokal),
         År=if_else(is.na(År),unique(År[!is.na(År)])[1],År)) %>%
  ungroup()
```

```{r}

# re-arrange effort data
effort_ind <- effort %>% filter(FISKE_ID %in% unique(Kul_data$FISKE_ID)) # only include cases where FISKE_ID is included in individual data

join_age_per_obsid_effort <- age_per_obsid %>% 
                             right_join(effort_ind,by=c("Lokal","År","OBS_ID","Djupstratum")) %>%  # we want to keep all net-night observations
                             complete(ind_age, nesting(OBS_ID, Djupstratum), fill=list(N=0)) %>% # include also zero catches(i.e. nets with no perch)
                             group_by(OBS_ID) %>% 
                             mutate(Lokal=if_else(is.na(Lokal),unique(Lokal[!is.na(Lokal)])[1],Lokal)) %>%#,
                        #            År=if_else(is.na(År),unique(År[!is.na(År)])[1],År)) %>%
                             ungroup() %>% 
                             filter(!is.na(ind_age),!is.na(År)) %>% # we don´t want na:s produced by joining
                             pivot_wider(id_cols=c(Lokal, Djupstratum, År, OBS_ID),names_from=ind_age,values_from=N)

```

### 3.2.5 Define minimum age and plus-group for each area

Here we define which age-classes to use in the model. The minimum age-class is the age-class at which mean CPUE per age-class is at its peak, and plus-groups are defined as the smallest age at which an age-class has a CPUE value larged than 0.2.

```{r}
ages_in_model <- 
join_age_per_obsid_effort %>%
  group_by(Lokal) %>%
  summarize(across(`1`:`10`, ~mean(.x, na.rm = TRUE))) %>%
  ungroup() %>%
  pivot_longer(cols=`1`:`10`,names_to = "age_class", values_to = "CPUE") %>%
  mutate(is_plus_group=CPUE<0.2) %>%
  group_by(Lokal) %>%
  summarize(plus_age=min(as.numeric(age_class)[is_plus_group])-1,
            min_age=age_class[which.max(CPUE)])

```

## 3.3 Define scenarios to run

Here we define how data should be subsetted. We divde data based on area and the depth strata at which multimesh gillnets were set.

```{r}

combinations <- join_age_per_obsid_effort %>%
                distinct(Lokal, Djupstratum) %>%
                right_join(ages_in_model, by=join_by("Lokal"))

```

# 4. Fit models

## 4.1 Compile model

```{r}
# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')
```

## 4.2 Fit all models

Here we fit models to all combinations of data subsets defined above, that is, for each site and depth stratum we fit separate models. This code chunk is slow!! Hence, only use once and load MCMC data from the repository (MCMCOutput/Sensitivity_analysis) for analysis.

```{r}
#| eval: false

for(c(i:nrow(combinations))) {
  
# Rearrange data from one area into columns showing age classes and a plusgroup
data_wide <- summarize_plusgroup(min_age = combinations$min_age[i], 
                    plus_age = combinations$plus_age[i], 
                    area = combinations$Lokal[i], 
                    Depth_strata = combinations$Djupstratum[i],
                    data = join_age_per_obsid_effort)

# Create a list for fitting Bayesian model
STAN_input <- create_STAN_input_list(data_wide)

success <- FALSE

while (!success) {
    
# fit model
fit <- model$sample(
  data = STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(STAN_input,chains=4),
  adapt_delta=0.95,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# save MCMC data
saveRDS(STAN_input, 
        paste0("../MCMCOutput/Sensitivity_analysis/",
                       "inputdata_",
                       combinations$Lokal[i],
                       "_",
                       combinations$Djupstratum[i],
                       ".RDS"))
        
fit$save_object(paste0("../MCMCOutput/Sensitivity_analysis/",
                       "fit_",
                       combinations$Lokal[i],
                       "_",
                       combinations$Djupstratum[i],
                       ".RDS"))

}
```

## 4.3 Load fitted data

Here we load data for all models based on various data subsets(Areas and depth strata). We load both MCMC results from the baseline models, i.e. models including all data, and models based on data subset. Thereafter, we evaluate what data subsetting does for the analysis.

```{r}
##############################################
# File paths to models based on data subsets #
##############################################
sensitivity_models_all_files <- list.files("../MCMCOutput/Sensitivity_analysis/",full.names="TRUE") # paths to saved files
sensitivity_models_fit_files <- grep("fit_", sensitivity_models_all_files, value = TRUE) # paths to mcmc model fit summary files
sensitivity_models_inputdata_files <- grep("inputdata_", sensitivity_models_all_files, value = TRUE) # paths to input data files

#####################################
# File paths to baseline model fits #
#####################################
baseline_models_all_files <- list.files("../MCMCOutput/",pattern=".RDS$",full.names="TRUE") # paths to saved files
baseline_models_fit_files <- grep("fit_", baseline_models_all_files, value = TRUE) # paths to model fits
baseline_models_inputdata_files <- grep("inputdata_", baseline_models_all_files, value = TRUE) # paths to input data files

# initialize list-vectors for saving model inputs and model outputs
sensitivity_model_fits <- vector("list",length(sensitivity_models_fit_files))
baseline_model_fits <- vector("list", length(baseline_models_fit_files))

# loop over all model fits to data subsets
for(i in c(1:length(sensitivity_models_fit_files))) {
  fit <- readRDS(sensitivity_models_fit_files[i]) # MCMC data
  input <- readRDS(sensitivity_models_inputdata_files[i]) # input data
  fit_summary <- restruct_STAN_fit(fit, input) # summarize MCMC fit
  Lokal <- sub(".*fit_([^_]+)_.*", "\\1", sensitivity_models_fit_files[i]) # Extract name of fitted area
  Depth_stratum <- sub(".*_([^_]+)\\.RDS$", "\\1", sensitivity_models_fit_files[i]) # Extract depth strata from file name
  
  # save a list with scenarios indices(Area and depth strata), model fit and input data. 
  sensitivity_model_fits[[i]] <- list(Area=Lokal, 
                                      Depth_stratum=Depth_stratum, 
                                      fit = fit_summary, 
                                      input = input)
} # end of for-loop

# loop over all baseline models
for(i in c(1:length(baseline_models_fit_files))){
  fit <- readRDS(baseline_models_fit_files[i]) # Load MCMC model fit data
  input <- readRDS(baseline_models_inputdata_files[i]) # Load input data
  fit_summary <- restruct_STAN_fit(fit, input) # Summarize MCMC data
  Lokal <- sub("fit_(.*)\\d{4}-\\d{2}-\\d{2}.*", "\\1", basename(baseline_models_fit_files[i])) # Extract name of fitted area
  
  # save a list with area, model fit and input data.
  baseline_model_fits[[i]] <- list(Area = Lokal,
                               Depth_stratum = "All Depths",
                               fit = fit_summary,
                               input = input) 
} # end of for-loop


# convert to tibble with first two columns representing scenarios
sensitivity_model_fits <- map_dfr(sensitivity_model_fits, ~ tibble(
  Area = .x$Area,
  Depth_stratum = .x$Depth_stratum,
  fit = list(.x$fit),
  input = list(.x$input)
)) %>% # rename a few areas
  mutate(Area=case_when(Area=="Finbo, Åland" ~ "Finbo",
                        Area=="Kumlinge, Åland" ~ "Kumlinge",
                        Area=="Torhamn, Karlskrona Ö skärgård" ~ "Torhamn",
                        .default = Area))

# convert to tibble with first two columns representing scenarios(in this case Area and depth stratum="All Depths")
baseline_model_fits <- map_dfr(baseline_model_fits, ~ tibble(
  Area = .x$Area,
  Depth_stratum = .x$Depth_stratum,
  fit = list(.x$fit),
  input = list(.x$input)
))

# Merge model summaries for baseline and depth stratified model fits
all_model_fits <- bind_rows(sensitivity_model_fits, baseline_model_fits)
```

# 5. Figures

## 5.1 Extract data needed for creating figures

```{r}
##################################
# extract data needed  for plots #
##################################

# 1. Mean mortality per age-class
Plot_data_Z_hat <- all_model_fits %>% #all_model_fits %>%
  mutate(selected = map(fit, "Z_hat")) %>%  # extract sublist
  unnest(selected) %>%
  filter(age_class!="plusgroup")

# 2. Time varying mortality per age-class
Plot_data_Z <- all_model_fits %>%
  mutate(selected = map(fit, "Z")) %>%  # extract sublist
  unnest(selected) %>%
  filter(age_class!="plusgroup")

# 3. Abundance weighted mortality
Plot_data_Z_mean <- all_model_fits %>%
  mutate(selected = map(fit, "Z_mean")) %>%  # extract sublist
  unnest(selected) 

# set wong-colour palette
pal <- set_wong_colours()

```

## 5.2 Plot figures

### 5.2.1 Mean age-class mortality

Mean age-class mortality was calculated as a mean mortality for the full time series, for each age-class individually:

$$
\bar{Z}_a = \frac{\sum_{t=1}^{T} \hat{Z}_{a,t}}{T}
$$

where $\hat{Z}_{a,t}}$ is the mortality rate for age-class a at time *t* and *T* is the time series length.

```{r}
#| label: fig-mean-age-class-mortality   
#| fig-cap: "Mean age-class mortality. This figure shows how model estimate of mean age-class mortality changes depending on data subsetting. Each bar represents mean age-class mortality estimates based on gillnet data from the depth stratum 0,1-3 m, 3-6 m, 6-10 m and all depths are here shown in yellow, light blue green and dark blue, respectively. Trendlines based on mean age-class mortality as a function of age-class are displayed in the same colours as bars."

# create figure  
  ggplot(Plot_data_Z_hat, aes(x = age_class, y = median, fill = Depth_stratum, group=Depth_stratum,colour=Depth_stratum)) +
  geom_col(position=position_dodge(width=0.8), width=0.7) +
  geom_smooth(method="lm", se=FALSE) +  
    labs(x = "Age class",
         y = expression(bar(italic(Z))[a] ~ "(" * "Mean age-class mortality" * ")")) +
    scale_fill_manual(name = "Depth stratum",
                      values = pal) +
    scale_colour_manual(name = "Depth stratum",
                        values = pal) +
    facet_wrap(~Area, scales="free_y") +
    theme_bw()

```

### 5.2.2 Timevarying age-class mortality

Time varying age-class-specific mortality rates were simply derived from model estimates:

$$
\hat{Z}_{a,t}}
$$

```{r}
#| label: fig-age-class-mortality 
#| fig-cap: "Age-class specific mortality rates. This figure shows how model estimate of time varying age-class specific mortality rates change depending on data subsetting. Mortality estimates based on gillnet data from the depth stratum 0,1-3 m, 3-6 m, 6-10 m and all depths are here shown in yellow, light blue green and dark blue, respectively."

# Creat a list of areas  
Areas <- unique(all_model_fits$Area)

# Function for plotting age-class mortality per area
age_class_mortality_per_depth <- function(Plot_data_Z, Area_name) {  
  ggplot(Plot_data_Z %>% filter(Area==Area_name), aes(x = year, y = median, color = Depth_stratum)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = expression(paste("Z"['a, t']," (Age-class mortality)")),
    title = Area_name
  ) +
  scale_color_manual(values = pal) +
  facet_wrap(vars(age_class),scales="free_y") +
  theme_bw()  
}

# loop over all areas and plot mortality per-age class and depth
for (i in 1:length(Areas)) {
  print(age_class_mortality_per_depth(Plot_data_Z, Areas[i]))
}

```

### 5.2.3 Abundance weighted mortality

Annual mortality rates were derived using abundance weighted age-class specific mortality rates, following:

$$
\bar{Z}_t = \frac{\sum_{i} \hat{Z}_{i,t} \cdot N_{i,t}}{\sum_{i} N_{i,t}}
$$

Annual age-class mortality is the posterior draws of the annual age-class mortality ($\hat{Z}[a,t]$) defined above.

```{r}
#| label: fig-abundance-weighted-mortality 
#| fig-cap: "Abundance weighted mortality. This figure shows how model estimates of abundance weighted mean mortality change depending on data subsetting. Mortality estimates based on gillnet data from the depth stratum 0,1-3 m, 3-6 m, 6-10 m and all depths are here shown in yellow, light blue green and dark blue, respectively."

  ggplot(Plot_data_Z_mean, aes(x = year, y = median, color=Depth_stratum)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
         y = expression(bar(italic(Z))[t] ~ "(" * "Abundance weighted mortality" * ")")
  ) +
  scale_color_manual(values = pal) +
  facet_wrap(vars(Area),scales="free_y") +
  theme_bw()  

```

# 6. Conclusion

-   The increasing mortality with age as found across all studied sites is robust against data subsetting. In other words, mortality increase with age if data from shallow stations, intermediate depth stations as well as deep stations are used when fitting models (@fig-mean-age-class-mortality). The pattern is, however, a bit less pronounced when data from shallow stations are use in the analysis. Nothwithstanding, the observation of an increasing mortality with age is still apparent also for the shallow station case.

-   In general, the temporal dynamics in age-class mortality (see @fig-age-class-mortality) is similar, with peaks and valleys in age-class specific mortality often co-occuring across data subsets. However, it should be noted that some discrepancies from this general pattern are apparent (see @fig-age-class-mortality).

-   The general mortality experienced by perch tend to be relatively robust to data sub setting, and mean mortality estimates tend to co-vary across the data subsets (@fig-abundance-weighted-mortality).

-   Overall, this analysis suggests that observed pattern of an increasing mortality with age is not an effect of perch utilizing different depth strata within areas. However, it might very well be so that perch utilize different bays during ontogeny, and that some fish disperse.
