---
title: "Lännåkersviken_Askviken"
author: "Torbjörn Säterberg"
date: today  
format: html
embed-resources: true
self-contained: true
editor: visual
code-fold: true
error: FALSE
warning: FALSE
message: FALSE
toc: true
toc-depth: 3
lang: sv  
---

## Introduction

Here we try to fit perch models for Lännåkersviken and Askviken. Age information is only available for 2 years(2009 and 2013). Hence, we pooled the information for these years and both areas. ALK was based on these data sets.

## 1.2 Required R-packages {#R-packages}

```{r}
#| label: load-packages
require("RPostgres", quietly = TRUE) # postgres functionality in R
library(DBI) # For database connection
library(odbc) # Connection between R and relation database management systems
library(ggforce, quietly=TRUE)  # added functionality to ggplot
library(tidybayes) # working with posterior draws for bayesian models
library(flextable) # plotting nice tables
library(brms) # multilevel models in a bayesian context(similar to lmer)
library(bayesplot) # plotting bayesian models
library(future) # parallell computing 
library(purrr) # working with functions in pipe
library(modelr) # modeling functions with pipe
library(tidyverse) # tidyverse functionality for working with data
library(cmdstanr) # fitting stan models through R
```

## 1.3 Connect to KUL-database {#dbconnection}

```{r}
# Connect to Aqua-datalagret
AQ_con <- dbConnect(odbc(), 
                         Driver = "SQL Server", 
                         Server = "aquadw_prod.db.slu.se",
                         USER = Sys.info()["user"], 
                         Database = "AquaDW_prod", 
                         Trusted_Connection = "TRUE", 
                         encoding = "UTF-8")
```

## 2.1 Load effort data {#effordata}

Here we load and summarize effort data from the KUL table "KUL_OBSERVATION_NEW". For perch we are only interested in gill-net nights that are undisturbed (störing=NEJ), approved (Godkänd=JA) and are sampled at 0-10 m depth.

```{sql, connection="AQ_con", output.var = "effort"}
-- CALCULATE EFFORT
SELECT Lokal, Fångstområde, År, FISKE_ID, OBS_ID
FROM AquaDW_prod.mart.KUL_OBSERVATION_NEW
WHERE
Redskap_kod = 'K064' 
AND
Störning='NEJ'
AND 
Godkänd ='JA'
AND
Fångstområde IN('Askviken','Lännåkersviken');
```

## 2.2 Load catch data

```{sql, connection="AQ_con", output.var = "Kul_data"}
SELECT *
FROM AquaDW_prod.mart.KUL_LANGDGRUPP klmt --
WHERE 
Redskap_kod = 'K064' 
AND
Art = 'Abborre' 
AND
Godkänd='JA'
AND 
Störning='NEJ'
AND
Fångstområde IN('Askviken','Lännåkersviken');
```

## 2.3 Load individual data

```{sql, connection="AQ_con", output.var = "Kul_ind"}
SELECT * 
FROM
AquaDW_prod.mart.KUL_INDIVID_NEW
WHERE 
Redskap_kod = 'K064' 
AND
Art = 'Abborre' 
AND
Fångstområde IN('Askviken','Lännåkersviken');

```

## 2.4 Load sex data

```{sql, connection="AQ_con", output.var = "Kul_sex"}
SELECT Lokal, Fångstområde, Fiskeår, Längdgrupp, ANTAL_HANAR, ANTAL_HONOR, ANTAL
FROM AquaDW_prod.mart.KUL_STICKPROVSUTFALL ks
WHERE
Art = 'Abborre' 
AND
Fångstområde IN('Askviken','Lännåkersviken');

```

## 3.1 Remove NAs from individual and catch data sets {#filterdata}

```{r}
Kul_data <- Kul_data %>% 
            filter(!is.na(Längdgrupp),!is.na(Antal)) %>% # remove cases with no length info
            uncount(Antal, .remove = FALSE) # expand df such that each row is one individual

Kul_ind <- Kul_ind %>% 
           filter(!is.na(Slutlängd_cm),!is.na(Ålder)) # we cant use NAs
```

## 3.2 Estimate age based indices using ALK

In this section we:

1.  [Derive age-length keys](#ALKs)([ALK](#ALK-function)) using [age-read individuals](#inddata)
2.  [Estimate age of individual fish](#estfishage) using [ALKs](#ALKs) and the [full data set with fish lengths](#catchdata).
3.  [Summarize catches per age class](#fishperlengthclass)
4.  [Restructure the summarized catches (3) per age into cohorts](#cohort).

### 3.2.1 ALK function {#ALK-function}

This function is used for deriving age-length keys for data subsets. It takes as input the observed lengths in [the total catch](#catchdata) (*lengthD*) and [age and length data](#inddata) from a subsample of the total catch to derive an age-length key (i.e. the probability of age given length). This is sometimes called a forward age-length key [(Ailloud & Hoenig 2019 ICES J Sci)](https://academic.oup.com/icesjms/article/76/6/1515/5480389).

```{r}
applyALK <- function(lengthD,indD){

# Find number of individuals per length and age in individual data
LADat <- xtabs(~Slutlängd_cm + Ålder, data=indD) # number of observations per age- and length-class 
ind_lidx <- sort(unique(indD$Slutlängd_cm)) # length groups in individual data
L_lidx <- sort(unique(lengthD$Längdgrupp)) # Length groups in length data

# initialize Length/age group matrix which will be fille later
l_agr_mat <- matrix(NA,length(L_lidx),dim(LADat)[2])
rownames(l_agr_mat) <- L_lidx 
colnames(l_agr_mat) <- colnames(LADat)
l_agr_mat[L_lidx %in% ind_lidx,] <- LADat[rownames(l_agr_mat)[L_lidx %in% ind_lidx],]
rowidNA <- as.numeric(rownames(l_agr_mat)[apply(l_agr_mat, 1, anyNA)]) # columns to fill with data

# Fill in values for length groups with no individual data (borrow age data from lower or higher length-classes)
for(i in rowidNA){
  if(i<min(ind_lidx)){
  l_agr_mat[rownames(l_agr_mat)==as.character(i),] <- LADat[as.character(min(ind_lidx)),] # take the smallest length group from ind data if Length observations are smaller than individual data
  } else if (i>max(ind_lidx)) {
  l_agr_mat[rownames(l_agr_mat)==as.character(i),] <- LADat[as.character(max(ind_lidx)),] # take age dist from the largest length group from ind data if Length observations are larger than individual data 
  }else{    
  # take the sum of age distributions for closest length groups if length group is not part of individual data 
  l_agr_mat[rownames(l_agr_mat)==as.character(i),] <- colSums(LADat[c(as.character(min(ind_lidx[ind_lidx>i])), 
                                                          as.character(max(ind_lidx[ind_lidx<i]))),])
  }
  }

# convert from number of ind per age- and length class to proportion of age-classes per length-class in sample.
ALK <- prop.table(l_agr_mat,margin=1)
return(ALK)
}
```

### 3.2.2 Derive ALKs for each subset of data (FISKE_ID) {#ALKs}

Age-length keys are here derived for each subset of data(FISKE_ID) using data of [indivual age-read and length measured fish](#inddata) in combination with the [full data set](#catchdata). The full data set is only used to give information on the number of length-classes needed in the age-length keys.

```{r}

# Apply function to grouped dataframes
ALK_all <- applyALK(Kul_data,Kul_ind)
```

### 3.2.3 Estimate age of individual fish {#estfishage}

Here we use ALK and the full data set, subsetted by FISKE_ID, to estimate age of individual fish in the full length data set.

```{r}
# For reproducibility set seed
set.seed(123)

# sample individual age across all individuals (assuming only females in the sample)
ind_age <- FSA::alkIndivAge(ALK_all,~Längdgrupp,data=Kul_data) %>%
  select(Fångstområde, År, OBS_ID, Längdgrupp, age)

```

## 3.3 Assign sex to individuals

1.  First we calculate the [proportion of females](#female_prop_area) per area from annual sub-samples.
2.  Second, we [randomly assign sex to individuals](#assignsex) based on the calculated proportion of females per area and length group.

### 3.3.1 Proportion females per area (i.e. aggregate data across years for each area) {#female_prop_area}

Here we aggregate data across years in order to calculate the proportion of females per length group and area.

```{r}

# female proportions per area and year
female_prop_Y <- Kul_sex %>% 
  mutate(females = case_when(ANTAL_HANAR==ANTAL ~ 0, # Number of males equal to total -> 0 females
                             !is.na(ANTAL_HONOR) ~ ANTAL_HONOR, # Number of females reported
                             .default = NA),
         total = case_when(!is.na(ANTAL) ~ ANTAL, # Total sample reported
                           (is.na(ANTAL)) & 
                           (!is.na(ANTAL_HANAR) & !is.na(ANTAL_HONOR)) 
                           ~ ANTAL_HANAR+ANTAL_HONOR,# Females and males reported -> prop can be calculated
                           .default=NA),
         female_prop=females/total # proportion females in size class from sample
                                  )

# femle proportions per area
female_prop_area <- 
  female_prop_Y %>% 
  filter(!if_any(c("females","total"),is.na)) %>% # remove cases where either females or the total number individuals in samples is na
  group_by(Längdgrupp) %>%
  summarize(total = sum(total), 
            females = sum(females),
            prop_females = sum(females)/sum(total)) %>%
  ungroup()

```

### 3.3.2 Randomly assign sex to indivuals {#assignsex}

Here we randomly assign sex to individuals using a bernoulli model:

$$
X_{ijkl} \sim \text{Bern}(p_{jk})
$$

where $X_{ijk}$ is a binary random variable and ($p_{jk}$) is the probability that an individual perch in area *j* and age class *k* is a female. $p_{jk}$ is estimated ($\hat{p}_{jk}$) using [sex proportions](#female_prop_area) observed in samples of [sex](#sexdata).

```{r}
# for reproducibility set seed
set.seed(123)

# join sex probabilities to individual data
ind_sex <- ind_age %>%
  left_join(female_prop_area,by=c("Längdgrupp"))

# randomly sample sex based on sex probabilities per length group
ind_sex$isfemale <- rbinom(n=nrow(ind_sex),size=1,p=ind_sex$prop_females)

# The previous line produces a few na:s for age groups that do not occur in the sex data. For these lines assign females to large individuals
ind_sex$isfemale[is.na(ind_sex$isfemale) & (ind_sex$Längdgrupp>30)] <- 1

# for small ones randomly sample sex based on 50-50 chance of male-female
ind_sex$isfemale[is.na(ind_sex$isfemale)] <- rbinom(n=sum(is.na(ind_sex$isfemale)),size=1,p=0.5) 
  
```

### 3.2.4 Summarize number of fish per age-class {#fishperageclass}

```{r}
age_per_obsid <- ind_sex %>%
  group_by(Fångstområde,År,OBS_ID,age) %>% 
  summarize(N=sum(isfemale)) %>% # number of females per age and net
  ungroup() %>%
  complete(OBS_ID, age, fill=list(N=0)) %>%
  group_by(OBS_ID) %>% 
  mutate(Fångstområde=if_else(is.na(Fångstområde),unique(Fångstområde[!is.na(Fångstområde)])[1],Fångstområde),
         År=if_else(is.na(År),unique(År[!is.na(År)])[1],År)) %>%
  ungroup()
```

```{r}

# re-arrange effort data
effort_ind <- effort %>% filter(FISKE_ID %in% unique(Kul_data$FISKE_ID)) # only include cases where FISKE_ID is included in individual data

join_age_per_obsid_effort <- age_per_obsid %>% 
                             right_join(effort_ind,by=c("Fångstområde","År","OBS_ID")) %>%  # we want to keep all net-night observations
                             complete(OBS_ID, age, fill=list(N=0)) %>% # include also zero catches(i.e. nets with no perch)
                             group_by(OBS_ID) %>% 
                             mutate(Fångstområde=if_else(is.na(Fångstområde),unique(Fångstområde[!is.na(Fångstområde)])[1],Fångstområde)) %>%#,
                        #            År=if_else(is.na(År),unique(År[!is.na(År)])[1],År)) %>%
                             ungroup() %>% 
                             filter(!is.na(age),!is.na(År)) %>% # we don´t want na:s produced by joining
                             pivot_wider(id_cols=c(Fångstområde, År, OBS_ID),names_from=age,values_from=N)

```

We want to find the peak in catch-curves here (maximum CPUE per age-class) for all areas. Also, if CPUE is lower than 0.2 for a given age-class it is considered to belong to the plusgroup. More specifically, an age-class is considered to belong to the plusgroup if its cpue is \< 0.2, alternatively if its the age-class just before the first age-class with a cpue \< 0.2. This is done in order to avoid fitting models with too little data at hand.

```{r}
ages_in_model <- 
join_age_per_obsid_effort %>%
  group_by(Fångstområde) %>%
  summarize(across(`1`:`10`, ~mean(.x, na.rm = TRUE))) %>%
  ungroup() %>%
  pivot_longer(cols=`1`:`10`,names_to = "age_class", values_to = "CPUE") %>%
  mutate(is_plus_group=CPUE<0.2) %>%
  group_by(Fångstområde) %>%
  summarize(plus_age=min(as.numeric(age_class)[is_plus_group])-1,
            min_age=age_class[which.max(CPUE)])

ft_ages_in_model <- autofit(flextable(ages_in_model))

# print nice looking table
ft_ages_in_model

```

# 4 Utility functions

Here are a number of functions that are implement that are needed for plotting results for each area. Inputs and outputs for each function is described in the code chunk below.

```{r}

############
## Figures #
############

#-------------------------------#
# proportion of zeros in sample #
#-------------------------------#
prop_zero <- function(x) mean(x == 0)

#-----------------------------#
# Posterior predictive checks #
#-----------------------------#
# INPUTS
# input_data : A list with data according to the data structure in age_model_STAN.stan
# fit : A cmdstan model fit from the age_model_STAN.stan
# n_rep : number of simulated posterior samples to use in figures
#
# OUTPUT
# Posterior predictive checks checking:
# 1. Density plots (simulated and observed number of individuals caught per net) per age - group
# 2. Proportion of zeros, i.e. how many zeros are in simulated and observed data per age - group
create_PPC_plots <- function(input_data, fit, n_rep, plottitle){

# inits  
post <- head(posterior::as_draws_matrix(fit$draws()),n_rep) # reorder posterior for plotting
A <- input_data$A # number of age-groups

for(i in 1:A){
    
    string <- paste0("^yrep.*,",i,"]$")
    col_ind <- grep(string, colnames(post), value=TRUE)
    
    # distribution of number of individuals caught and n_rep prediction of these
    p1 <- ppc_dens_overlay(y = input_data$y[,i],
                     yrep = post[,col_ind]) +
             labs(title = paste0("Density age-group ",i," ",plottitle))
    
    p2 <- ppc_stat(y=input_data$y[,i], 
             yrep = post[,col_ind], 
             stat = "prop_zero", 
             binwidth = 0.005) +
             labs(title = paste0("Proportion zeros age-group ",i," ",plottitle))
    
    print(p1)
    print(p2)
  }
}

visualize_pred_and_obs_per_age_class <- function(STAN_fit_summary, data_long, plottitle){

# Time series plot for different age-classes
data_long %>% 
    group_by(year,age_class) %>% # first calculate summary statistic for observations
    summarize(q5=quantile(n,probs=0.05,type=3), # 5 percent quantile
              median=quantile(n,probs=0.5,type=3), # median
              q95=quantile(n,probs=0.95,type=3)) %>% # 95 percent quantile
    ungroup() %>%
ggplot(.,aes(x=year-0.15, y=median, ymin=q5, ymax=q95)) + # plot median and quantiles for observations
  geom_pointrange(color="black",linetype="dashed") +
  labs(
  x = "Year",
  y = expression(tilde(y) ~ "[a, t] (Number of individuals caught per net-night)")
  ) +
  ggtitle(plottitle) +  
  geom_pointrange(data=STAN_fit_summary$N_obs,color="red",aes(x=year+0.15, y=median, ymin=q5, ymax=q95)) + # plot predictions 
  facet_wrap(vars(age_class),scales="free_y")
}

visualize_expected_CPUE_per_age_class <- function(STAN_fit_summary, data_long, plottitle){
  # Time series plot for different age-classes
data_long %>% 
  group_by(year,age_class) %>% # first calculate summary statistic for observations
  summarize(mean_CPUE=mean(n)) %>% # calculate mean cpue
  ungroup() %>%
  ggplot(.,aes(x=year, y=mean_CPUE)) + # plot median and quantiles for observations
  geom_point(color="black") +
  labs(
    x = "Year",
    y = expression(bold(N["a,t"]))
  ) +
  ggtitle(plottitle) +  
  geom_line(data = STAN_fit_summary$N, aes(x = year, y = median), colour="red") +
  geom_point(data = STAN_fit_summary$N, aes(x = year, y = median), colour="red") +
  geom_ribbon(data = STAN_fit_summary$N,aes(x = year, y = median, ymin = q5, ymax = q95), alpha = 0.3, fill="red") +
  facet_grid(vars(age_class),scales="free_y")
}  
  
visualize_cohorts <- function(STAN_fit_summary, data_long, plottitle){

# first calculate median values per cohort of raw data  
median_per_year_and_year_class <-     
data_long %>% 
    group_by(year, year_class) %>% 
    summarize(median=median(n)) %>% # calculate median per year and year-class(this is what´s being predicted by the model) 
    ungroup() 

nb_year_classes <- length(unique(data_long$year_class))

# make a plot  
ggplot(STAN_fit_summary$N, aes(x=year, y=median, group=as.factor(year_class), colour=as.factor(year_class))) +
  geom_point(data = median_per_year_and_year_class, aes(shape = as.factor(year_class))) +
  geom_line(aes(linetype = as.factor(year_class))) +
  scale_shape_manual(values = 1:nb_year_classes, name = "Observations(median)") +
  scale_linetype_manual(values = 1:nb_year_classes, name = "Year-class predictions") +
  scale_color_viridis_d(option = "magma", name = "Year Classes",guide="none") +
  labs(y = expression(tilde(y) ~ "[a, t] (Number of individuals caught per net-night)"),
       title = plottitle) +
  theme(
      legend.position = "right", # Position legend on the right
      legend.box = "vertical",  # Arrange legends vertically
      legend.spacing.y = unit(0.1, "cm"), # Increase spacing between legend elements
      legend.key.size = unit(0.75, "lines"),  # Reduce size of legend keys
      legend.text = element_text(size = 8) # Reduce font size of legend text
  ) +  
  guides(
      linetype = guide_legend(order = 1),
      shape = guide_legend(order = 2)
  )
  }

visualize_annual_mortality <- function(STAN_fit_summary, plottitle) {
ggplot(STAN_fit_summary$f, aes(x = year, y = median)) +
  geom_line(colour="red") +
  geom_point(colour="red") +
  geom_ribbon(aes(ymin = q5, ymax = q95), alpha = 0.3, fill="red") +
  labs(
    x = "Year",
    y = expression(paste(f[t]," (Annual mortality)")),
    title = plottitle
  ) 
}

visualize_age_class_annual_mortality <- function(STAN_fit_summary, plottitle) {
ggplot(STAN_fit_summary$Z, aes(x = year, y = median)) +
  geom_line(colour="red") +
  geom_point(colour="red") +
  geom_ribbon(aes(ymin = q5, ymax = q95), alpha = 0.3, fill="red") +
  labs(
    x = "Year",
    y = expression(paste("Z"['a, t']," (Age-class mortality)")),
    title = plottitle
  ) +
  facet_wrap(vars(age_class),scales="free_y")  
}

make_mortality_table <- function(STAN_fit_summary, plottitle){
data <- STAN_fit_summary$Z %>% 
  select(variable, year, age_class, median, q5, q95) %>%  # select parameters to include in output
  mutate(median=round(median,2), q5=round(q5,2), q95=round(q95,2)) # round values
ft <- flextable(data)
set_caption(ft,paste("Age-class mortality", plottitle))
}

make_mean_age_class_mortality_table <- function(STAN_fit_summary, plottitle){
data <- STAN_fit_summary$Z_hat %>% 
  select(age_class, median, q5, q95) %>%  # select parameters to include in output
  mutate(median=round(median,2), q5=round(q5,2), q95=round(q95,2)) # round values
ft <- flextable(data)
set_caption(ft,paste("Mean age-class mortality", plottitle))
}

visualize_year_class_mortality <- function(STAN_fit_summary, plottitle) {
ggplot(STAN_fit_summary$sA) +
  geom_bar( aes(x = age_class, y = median), stat="identity", fill="red", alpha=0.7) +
  geom_pointrange( aes(x = age_class, y = median, ymin = q5, ymax = q95), colour = "red") +
  xlab("Age-class") +
  ylab(expression(paste(Delta~a," (% difference to annual mortality)"))) +
  ggtitle(plottitle)  
}

visualize_mean_age_class_mortality <- function(STAN_fit_summary, plottitle) {
ggplot(STAN_fit_summary$Z_hat) +
  geom_bar( aes(x = age_class, y = median), stat="identity", fill="red", alpha=0.7) +
  geom_pointrange( aes(x = age_class, y = median, ymin = q5, ymax = q95), colour = "black") +
  xlab("Age-class") +
  ylab(expression("Mean age-class mortality " * (bar(Z)[a]))) +
  ggtitle(plottitle)  
}

visualize_recruitment_index <- function(STAN_fit_summary, plottitle) {
  # Recruitment index
ggplot(STAN_fit_summary$R, aes(x = year, y = median)) +
  geom_line(colour = "red") +
  geom_point(colour = "red") +
  geom_ribbon(aes(ymin = q5, ymax = q95), alpha = 0.3, fill= "red") +
  labs(
    x = "Year",
    y = expression(paste(R[year]," (Recruitment index)")),
    title = plottitle
  ) +
  scale_y_log10()

}

# Time series plot for total stock size
visualize_pred_and_obs_total_stock_size <- function(STAN_fit_summary, data_long, plottitle){

# Time series plot for total stock size
data_long %>% 
    group_by(year,OBS_ID) %>% 
    summarize(n=sum(n)) %>% 
    ungroup() %>% 
    group_by(year) %>% 
    summarize(obs=mean(n)) %>%
    ungroup() %>% 
    ggplot(., aes(x=year, y=obs)) +
    geom_point() +
    geom_line(data = STAN_fit_summary$N_tot, aes(x = year, y = median), colour="red") +
    geom_point(data = STAN_fit_summary$N_tot, aes(x = year, y = median), colour="red") +
    geom_ribbon(data = STAN_fit_summary$N_tot,aes(x = year, y = median, ymin = q5, ymax = q95), alpha = 0.3, fill="red") +
    labs(
    x = "Year",
    y = expression(paste(N[tot], " (Total stock size)")),
    title = plottitle
    )
    

}


###################
# Manipulate data #
###################

#------------------------------------------------#
# Summarize data into age-groups and a plusgroup #
#------------------------------------------------#
# INPUTS
# min_age - minimum age of fish recruiting to the gear
# plus_age - The age above which few individuals reach
# area - the area of interest
# data - A tibble with columns areas, years and age-classes, and rows 
#        representing the number of individuals caught per net-night.
#
# OUTPUT
# dat - A tibble with similar structure as data, but old age-classes have been 
#       summarized into one age-class representing number of individuals above
#       a certain age caught per net-night, and only one area has been filtered 
#       out.

summarize_plusgroup <- function(min_age, plus_age, area, data) {

if (sum(area %in% unique(data$Fångstområde))!=1){
  print("Error: Lokal(area) not in your data")
  return(NULL)
}  
  
# find maximum age(columnname) in data  
max_age <-  max(as.numeric(
    grep("^\\d+$", names(data), value = TRUE)
    ))

# Ages of plus group
pgroup <- as.character(c(plus_age:max_age))

# Ages of age groups considered
age_groups <- as.character(c(min_age:(plus_age-1)))  

# sum across all ages above plusgroup age
dat <- data %>%
  filter(Fångstområde==area) %>% # filter one area
  mutate(plusgroup=rowSums(across(matches(pgroup)))) %>% # sum across plusgroup ages
  select(Fångstområde, År, OBS_ID, all_of(age_groups), plusgroup) %>% # select some of the variables
  arrange(År) %>% # sort based on year
  rename(year=År) # rename to year

return(dat)
}

#--------------------------#
# Long format for plotting #
#--------------------------#
#
# Restructure data into long format for plotting
#
create_long_format <- function(data){
  
  # age groups
  age_groups <- grep("^\\d+$", names(data), value = TRUE)
  
  # minimum age of plusgroup
  plusage <- max(as.numeric(age_groups)) + 1
  
  # age of first age-class
  N1_age <- min(as.numeric(age_groups))
  
  # columns to pivot
  cols=c(age_groups,"plusgroup") # we also want data for plusgroup
  
  # pivot data
  data_long <- data %>%
    pivot_longer(cols = all_of(cols), names_to = "age_class", values_to = "n") %>%
    mutate(year_class = case_when(age_class == "plusgroup" ~ year - plusage,
                                  age_class != "plusgroup" ~ year - as.numeric(age_class))) 
  
  return(data_long)
}

#----------------------------------------------#
# Create list with data for fitting STAN-model #
#----------------------------------------------#
#
# Outputs a data list with variables used in the STAN model.
#
create_STAN_input_list <- function(data){
  
# number of age groups
A <- length(grep("^\\d+$", names(data), value = TRUE)) + 1 # all numeric columnnames(age-classes) and the plus group

# years in time series
years <- min(data$year):max(data$year) # this includes also years with no data

if(length(min(years):max(years))!=length(unique(data$year))){
 print("Be aware that there are years with missing data in the data set!!")
}

# number of years in time series
Y <- length(years)

# indicator variable showing which time indicies have no data
Y_missing <- as.integer(years %in% setdiff(years,unique(data$year)))

# number of years with missing data
n_missing <- sum(Y_missing)

# Effort(i.e. number of net-nights) 
E <- (data %>% group_by(year) %>% summarize(eff=n()))

E <- E %>% complete(year=years) %>% 
  mutate(eff=ifelse(is.na(eff),0,eff))

E <- E$eff

# number of caught individuals per net-night
age_groups <- grep("^\\d+$", names(data), value = TRUE) # all numeric columns
y <- data %>% select(all_of(age_groups),"plusgroup")

# total number of observations
n <- nrow(data)

# initial age-group in data
N1_age <- as.numeric(min(age_groups))

# mean on log recruitment(number of individuals caught for the youngest age-class)
mean_log_N1 <- data %>% 
           group_by(year) %>% 
           summarize(across(min(age_groups), mean, .names="init")) %>%
           ungroup() %>%
           select(init) %>%
           filter(init!=0) %>%
           summarize(mean_log_N1=mean(log(init)))

mean_log_N1 <- mean_log_N1$mean_log_N1

# prior mean for first year of observations
Na1_init <- data %>%
            filter(year==min(year)) %>%
            select(-Fångstområde,-year,-OBS_ID) %>%
            summarize_all(mean) %>% # mean at age for first year or data(if zero assign small value)
            unlist()

for(i in 1:length(Na1_init)){ # if some values zero set these to the mean value across the full data set
if(Na1_init[i]==0){
  Na1_init[i] <- mean(unlist(data[,names(Na1_init[i])]))
}  
}



            
STAN_input <- list(A=A, # number of age groups(including plusgroup)
                   Y=Y, # total number of years
                   Y_missing=Y_missing, # Indicator of years with missing data
                   n_missing=n_missing, # number of years with missing data
                   years=years, # years with data
                   E=E, # total effort per year
                   n=n, # total number of observations(i.e. net-nights)
                   N1_age=N1_age, # the youngst age-class considered for modelling
                   mean_log_N1=mean_log_N1, # mean number of individuals of the first age-class per net-night and year
                   Na1_init=Na1_init, # mean number of individuals caught per net-night the first survey year
                   y=as.matrix(y)) # number of individuals caught per age-class and net-night (ages-class in column and observation per row) 

return(STAN_input)  
}


#----------------------------------------#
# Generate initial values for STAN-model #
#----------------------------------------#
#
# Outputs initial values, based on priors, for all parameters in model.
#
generate_inits_list <- function(data, chains = 4) {
  inits_list <- vector("list", chains)
  
  for (i in seq_len(chains)) {
    inits_list[[i]] <- list(
      phi = rgamma(1, 2, 1),
      mulogN1 = rnorm(1, data$mean_log_N1, 1),
      sdlogN1 = abs(rcauchy(1, 0, 1)),
      N1 = rlnorm(data$Y, data$mean_log_N1, 1),
      Na1 = rlnorm(data$A, log(data$Na1_init), 1),
      logZ1 = rnorm(data$A, log(0.2), 1),
      phi_logZ_prior = rbeta(1, 2, 2),
      sd_logZ = abs(rnorm(data$A, 0, 1)),
      z_logZ = array(rnorm(data$A * (data$Y - 1), 0, 1), dim = c(data$A, data$Y - 1))
    )
  }
  
  return(inits_list)
}

#---------------------------#
# Restructure STAN fit data #
#---------------------------#
restruct_STAN_fit <- function(fit, STAN_input_list){

# Some variables from STAN_input_list that are needed
A <- STAN_input_list$A 
N1_age <- STAN_input_list$N1_age # initial age-class considered
age_class <- names(STAN_input_list$Na1_init) #character
year <- STAN_input_list$years
Y <- STAN_input_list$Y
plusage <- A + N1_age - 1

# N - expected number of individuals per age-class
N <- fit$summary("N") %>% 
  mutate(age_class = factor(str_extract(variable, "(?<=\\[)[0-9]+"), 
                              levels = 1:A,        # Extract row index
                              labels = age_class), # Assign age_class to row index 
    year = factor(str_extract(variable, "(?<=,)[0-9]+"), 
                  levels = 1:Y,  # Extract column index
                  labels = year)  # Assign years to column index
  ) %>%
  mutate(age_class=as.character(age_class), # change from factor to character
         year=as.integer(as.character(year))) %>% # change from factor to character
  mutate(year_class=case_when(age_class=="plusgroup" ~ year - plusage, # Assign year class
                              age_class!="plusgroup" ~ year - as.numeric(age_class))) %>% 
  relocate(year,year_class,age_class,.after=variable) # rearrange order of columns


# N - estimate of number of individuals per age-class
N_obs <- fit$summary("N_obs") %>% 
  mutate(age_class = factor(str_extract(variable, "(?<=\\[)[0-9]+"), 
                              levels = 1:A,        # Extract row index
                              labels = age_class), # Assign age_class to row index 
    year = factor(str_extract(variable, "(?<=,)[0-9]+"), 
                  levels = 1:Y,  # Extract column index
                  labels = year)  # Assign years to column index
  ) %>%
  mutate(age_class=as.character(age_class), # change from factor to character
         year=as.integer(as.character(year))) %>% # change from factor to character
  mutate(year_class=case_when(age_class=="plusgroup" ~ year - plusage, # Assign year class
                              age_class!="plusgroup" ~ year - as.numeric(age_class))) %>% 
  relocate(year,year_class,age_class,.after=variable) # rearrange order of columns

# N_tot - number of individuals per year
N_tot <- bind_cols(as.data.frame(year), fit$summary("N_tot")) %>%
  relocate(year,.after=variable)

# Recruitment indicies
R <- bind_cols(as.data.frame(year), fit$summary("R")) %>%
  relocate(year,.after=variable)

# Annual mortality per age-class
Z <- fit$summary("Z") %>% 
  mutate(age_class = factor(str_extract(variable, "(?<=\\[)[0-9]+"), 
                              levels = 1:A,        # Extract row index
                              labels = age_class), # Assign age_class to row index 
    year = factor(str_extract(variable, "(?<=,)[0-9]+"), 
                  levels = 1:(Y-1),  # Extract column index
                  labels = year[1:Y-1])  # Assign years to column index
  ) %>%
  mutate(age_class=as.character(age_class), # change from factor to character
         year=as.integer(as.character(year))) %>% 
  relocate(year,age_class,.after=variable) # rearrange order of columns

# mean mortality per age-class
Z_hat <-
fit$summary("Z_hat") %>% mutate(age_class = factor(str_extract(variable, "(?<=\\[)[0-9]+"), 
                                                   levels = 1:A,        # Extract row index
                                                   labels = age_class)) %>%
  mutate(age_class=as.character(age_class)) %>% 
  relocate(age_class,.after=variable) # rearrange order of columns


# put all parameters in a list
STAN_output <- list(N=N, # Expected number of individuals observed per age-class
                    N_obs=N_obs, # Estimate of number of individuals observed per net-night
                    N_tot = N_tot, # Estimated stock size (deterministic prediction)
                    R=R, # Estimated recruitment for 0-year olds (Note!!! 0+ individuals in august)
                    Z=Z, # Mortality per age-class
                    Z_hat=Z_hat) # Mean mortality per ageclass 
}
```

# 5 Fit models

## 5.1 Lännåkersvikent

Add a description of the area

### 5.1.1 Format data

```{r}
# Rearrange data from one area into columns showing age classes and a plusgroup
data_wide <- summarize_plusgroup(min_age = 1, 
                    plus_age = 8, 
                    area = "Lännåkersviken", 
                    data = join_age_per_obsid_effort)

# Rearrange into long format for plotting etc.
data_long <- create_long_format(data_wide)

# Create a list for fitting Bayesian model
STAN_input <- create_STAN_input_list(data_wide)

# Add title for plots for this area
plottitle = "Lännåkersviken"
```

### 5.1.2 Fit model

```{r}
#| include: false

# compile model
model <- cmdstan_model('../STAN/age_model_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

### 5.1.3 Model diagnostics

```{r}
##| eval: FALSE # This needs to be run for each area

# check for potential sampling issues
print(fit$cmdstan_diagnose()) # Model fit is ok!!

```

### 5.1.4 Model fit

```{r}

# Plot posterior predictive checks(densities per age-group and number of zeros per age-group)
create_PPC_plots(STAN_input, fit, 100, plottitle)

# calculate bayesian R2-value
R2 <- median(bayes_R2(posterior::as_draws_matrix(fit$draws("yrep")),c(STAN_input$y))) # need a matrix with each obs simulated in columns and and vector with observations as second argument.

```

### 5.1.5 Stock indicators

```{r}

# Make the fitted stan object suitable for plotting
fit_summary <- restruct_STAN_fit(fit, STAN_input)

# Time series plot for different age-classes
visualize_pred_and_obs_per_age_class(fit_summary, data_long, plottitle = plottitle)

# Time series plot for total stock size
visualize_pred_and_obs_total_stock_size(fit_summary, data_long, plottitle = plottitle)

# Plot expected CPUE per age-class
visualize_expected_CPUE_per_age_class(fit_summary, data_long, plottitle=plottitle)

# plot annual age-class mortality
visualize_age_class_annual_mortality(fit_summary, plottitle = plottitle)

# Make table with mortality estimates
make_mortality_table(fit_summary,plottitle = plottitle)

# Mean age-class mortality
visualize_mean_age_class_mortality(fit_summary, plottitle)

# Make table with mean age-class mortalities
make_mean_age_class_mortality_table(fit_summary,plottitle)

# Plot recruitment index(an estimate of the number of 0+ individuals at census [i.e. in august])  
visualize_recruitment_index(fit_summary, plottitle = plottitle)

# cohort plot
visualize_cohorts(fit_summary, data_long, plottitle = plottitle)

```

## 5.2 Askviken

Add a description of the area

### 5.2.1 Format data

```{r}
# Rearrange data from one area into columns showing age classes and a plusgroup
data_wide <- summarize_plusgroup(min_age = 3, 
                    plus_age = 8, 
                    area = "Askviken", 
                    data = join_age_per_obsid_effort)

# Rearrange into long format for plotting etc.
data_long <- create_long_format(data_wide)

# Create a list for fitting Bayesian model
STAN_input <- create_STAN_input_list(data_wide)

# Add title for plots for this area
plottitle = "Askviken"
```

### 5.2.2 Fit model

```{r}
#| include: false

# compile model
model <- cmdstan_model('../STAN/age_model_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

### 5.2.3 Model diagnostics

```{r}
##| eval: FALSE # This needs to be run for each area

# check for potential sampling issues
print(fit$cmdstan_diagnose()) # Model fit is ok!!

```

### 5.2.4 Model fit

```{r}

# Plot posterior predictive checks(densities per age-group and number of zeros per age-group)
create_PPC_plots(STAN_input, fit, 100, plottitle)

# calculate bayesian R2-value
R2 <- median(bayes_R2(posterior::as_draws_matrix(fit$draws("yrep")),c(STAN_input$y))) # need a matrix with each obs simulated in columns and and vector with observations as second argument.

```

### 5.2.5 Stock indicators

```{r}

# Make the fitted stan object suitable for plotting
fit_summary <- restruct_STAN_fit(fit, STAN_input)

# Time series plot for different age-classes
visualize_pred_and_obs_per_age_class(fit_summary, data_long, plottitle = plottitle)

# Time series plot for total stock size
visualize_pred_and_obs_total_stock_size(fit_summary, data_long, plottitle = plottitle)

# Plot expected CPUE per age-class
visualize_expected_CPUE_per_age_class(fit_summary, data_long, plottitle=plottitle)

# plot annual age-class mortality
visualize_age_class_annual_mortality(fit_summary, plottitle = plottitle)

# Make table with mortality estimates
make_mortality_table(fit_summary,plottitle = plottitle)

# Mean age-class mortality
visualize_mean_age_class_mortality(fit_summary, plottitle)

# Make table with mean age-class mortalities
make_mean_age_class_mortality_table(fit_summary,plottitle)

# Plot recruitment index(an estimate of the number of 0+ individuals at census [i.e. in august])  
visualize_recruitment_index(fit_summary, plottitle = plottitle)

# cohort plot
visualize_cohorts(fit_summary, data_long, plottitle = plottitle)

```

# 6. Comment/conclusion

After attempting to fit age-based models to data from Lännåkersviken, Askviken and their combined dynamics it was found that this is very difficult. Inconsistencies such as higher CPUE for a given age-class a+1 at time t+1 than CPUE for age-class a at time t were consistently found in this data set. This is problematic and may stem from the fact that we do not know to what extent ALK changes across years due to differences in growth. As growth rates have increased over time, as found in [Fiskbarometern](https://fiskbarometern.se/rapport/2024/species/Abborre), it may be erroneous to fit a model assuming a stable age-length structure. 
