---
title: "Simulation_testing_of_perch_model"
author: "Torbjörn Säterberg"
date: today  
format: html
embed-resources: true
self-contained: true
editor: visual
code-fold: true
error: FALSE
warning: FALSE
message: FALSE
toc: true
toc-depth: 4
lang: en  
---

# 1. Introduction

This script is used for testing the reliability of an age-based perch model developed by Säterberg & Olsson ([2026](https://pub.epsilon.slu.se/39154/1/saterberg-t-et-al-20260112.pdf)) using computer simulations.

We estimate bias in age-specific mortality rate estimates using computer [simulations](#computer_simulations) and are specifically interested in a potential bias in time and age-specific mortality rates, and a potential bias in mean mortality rate per age-class.

The analyses are done as we want to assess to what extent the model produces biased estimates. Importantly, as the general result from implementing the age-based model on coastal data suggests that mortality increases with perch age ([Säterberg & Olsson 2026](https://pub.epsilon.slu.se/39154/1/saterberg-t-et-al-20260112.pdf)), we want to address whether this could be due to a model artifact. For example, will a decreasing morality with age be recovered by the model or does the model always suggest an increasing mortality with age independent of the structure of the simulation.

# 1.1 R-packages

```{r}
library(MASS)  # Stats functions
library(tidyverse) # tidy functions
library(cmdstanr) 
```

# 1.2 Source utility functions and Simulation files

```{r}

# List all .R simulation files in the directory
files.sources <- list.files(path = "../R-simulations", pattern = "\\.R$", full.names = TRUE)
invisible(sapply(files.sources, source)) # source

# List utility functions
files <- list.files("../R", pattern = "\\.R$", full.names = TRUE)
invisible(sapply(files, source)) # source
 
```

# 2. Simulation based bias estimation {#computer_simulations}

In order to test the reliability of the perch model we simulated data, trying to mimic the general pattern observed in mortality rates observed in the applications of the model to coastal perch data. With this regards, the general pattern in mortality rates observed in applications to coastal perch data, that is, increasing mortality with age and trends in mortality over time, were simulated. Moreover, in order to simulate biologically plausible population dynamics, and mortality patterns, parameter values were set to the ranges observed in model applications to coastal perch data, parameter values were set inspired by Råneå (A system with 7 age classes) and Kinnbäcksfjärden (A system with 4 age classes) as example system.

## 2.1 Model overview

Similar to the statistical model, we simulated age-structured population dynamics using a discrete-time model with stochastic recruitment, time-varying mortality, and observation error.

The model tracks abundance-at-age $N_{a,t}$ for age classes $a = 1, \dots, A$ and time steps $t = 1, \dots, T$.

### 2.1.1 Recruitment

Recruitment to the first age class was modeled as a lognormal process:

$$
N_{1,t} \sim \text{Lognormal}(\mu_R, \sigma_R^2)
$$

where $\mu_R$ is the mean log recruitment and $\sigma_R$ is the standard deviation on the log scale.

### 2.1.2 Mortality dynamics

Total mortality $Z_{a,t}$ varies over time according to a deterministic signal plus stochastic noise.

#### 2.1.2.1 Deterministic signals

Two alternative temporal trends were implemented.

**Sinusoidal variation:**

$$
\log Z_t^{\text{signal}} = A \sin(2 \pi f t + \phi)
$$

```{r}
#| label: fig-sinus-signal
#| fig-cap: "Sinus signal implemented as a stationary mortality pattern"
#| fig-width: 6
#| fig-height: 4

#####################
# Plot sinus signal #
#####################

# Time
t <- seq(0,20,by=0.1)

# Sinus parameters
A <- 0.5      # amplitude
f <- 0.1      # frequency (cycles per time unit)
phi <- runif(1, 0, 2*pi)   # random phase shift

# sinus signal
signal <-  A * sin(2 * pi * f * t + phi)

# create data frame
data <- tibble(t, signal)

# plot sinus trend
ggplot(data, aes(x=t, y=signal)) + 
  geom_line() +
  theme_bw() +
  labs(x="Time", 
       y = expression(log(Z[t]^{"signal"})))

```

**Logistic trend:**

$$
\log Z_t^{\text{signal}} = \frac{K}{1 + \exp(-r (t - t_{\text{mid}}))}
$$

```{r}
#| label: fig-logistic-trend
#| fig-cap: "A logistic trend implemented as a case of an increasing mortality trend"
#| fig-width: 6
#| fig-height: 4

#####################
# Plot sinus signal #
#####################

# Time
t_max <- 20
t <- seq(0,t_max,by=0.1)

# logistic trend parameters  
K <- 1
r <- 0.5
t_mid <- (t_max-1)/2
signal <- K / (1 + exp(-r * (t - t_mid)))

# create data frame
data <- tibble(t, signal)

# plot logistic trend
ggplot(data, aes(x=t, y=signal)) + 
  geom_line() +
  theme_bw() +
  labs(x="Time", 
       y = expression(log(Z[t]^{"signal"})))

```

#### 2.1.2.2 Noise

Noise and average age class specific mortality rates was added to the deterministic mortality signal:

$$
\log Z_{a,t} = \mu_a + \log Z_t^{\text{signal}} + \epsilon_{a,t}
$$

where

$$
\epsilon_{a,t} \sim \mathcal{N}(0, \sigma_Z^2)
$$

The variance was scaled relative to the signal:

$$
\sigma_Z = \kappa \cdot \text{SD}(\log Z^{\text{signal}})
$$

Mortality was then transformed back to the natural scale:

$$
Z_{a,t} = \exp(\log Z_{a,t})
$$

### 2.1.3 Population dynamics

Abundance-at-age followed standard cohort dynamics.

For ages $a < A$:

$$
N_{a,t} = N_{a-1,t-1} \cdot \exp(-Z_{a-1,t-1})
$$

For the plus group ($a = A$):

$$
N_{A,t} = N_{A-1,t-1} \cdot \exp(-Z_{A-1,t-1}) + N_{A,t-1} \cdot \exp(-Z_{A,t-1})
$$

### 2.1.4 Observation model

Observed counts were generated using a negative binomial distribution:

$$
y_{i,a,t} \sim \text{NegBin}(N_{a,t}, \phi)
$$

where $i$ indexes sampling events (gillnet nights) and $\phi$ is the dispersion parameter.

Sampling effort was assumed constant within each year.

### 2.1.5 Simulation implementation

The simulation model was implemented in R and consisted of the following steps:

1.  Simulating recruitment
2.  Generating mortality trajectories
3.  Propagating population dynamics
4.  Adding observation error
5.  Formatting outputs for Stan

100 replicates of population dynamics, with t=20, 30 gillnet nights were simulated and thereafter assessed using both absolute and relative bias.

### 2.1.6 Simulation scenarios

| Scenario | Number of age-classes | Mortality-time-pattern | Mortality-age-relationship | Noise vs signal |
|---------------|---------------|---------------|---------------|---------------|
| S1 | 4 | Stationary | Increase with age | 0.5 |
| S2 | 4 | Stationary | Same for all age-classes | 0.5 |
| S3 | 4 | Stationary | Decrease with age | 0.5 |
| S4 | 4 | Stationary | Increase with age | 1 |
| S5 | 4 | Stationary | Same for all age-classes | 1 |
| S6 | 4 | Stationary | Decrease with age | 1 |
| S7 | 7 | Stationary | Increase with age | 0.5 |
| S8 | 7 | Stationary | Same for all age-classes | 0.5 |
| S9 | 7 | Stationary | Decrease with age | 0.5 |
| S10 | 7 | Stationary | Increase with age | 1 |
| S11 | 7 | Stationary | Same for all age-classes | 1 |
| S12 | 7 | Stationary | Decrease with age | 1 |
| S13 | 4 | Increasing | Increase with age | 0.5 |
| S14 | 4 | Increasing | Same for all age-classes | 0.5 |
| S15 | 4 | Increasing | Decrease with age | 0.5 |
| S16 | 4 | Increasing | Increase with age | 1 |
| S17 | 4 | Increasing | Same for all age-classes | 1 |
| S18 | 4 | Increasing | Decrease with age | 1 |
| S19 | 7 | Increasing | Increase with age | 0.5 |
| S20 | 7 | Increasing | Same for all age-classes | 0.5 |
| S21 | 7 | Increasing | Decrease with age | 0.5 |
| S22 | 7 | Increasing | Increase with age | 1 |
| S23 | 7 | Increasing | Same for all age-classes | 1 |
| S24 | 7 | Increasing | Decrease with age | 1 |

: Simulation scenarios {#tbl-sim}

# 3. Results

## 3.1 Stationary mortality

### 3.1.1 Scenario 1: 4 age-classes, mortality increases with age, mortality stationary over time and 0.5 noise to signal

#### 3.1.1.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
logSD = c(0.8, 0.5, 0.2),                                       # SD on the log scale of mortality variability 
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-2.3710604, -0.3879585,  0.4002884,  0.4002884),  # Mean Z on the log-scale
Z_CV = c(1.02, 0.51, 0.124),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 3.1.1.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario1.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario1.RDS")
```

#### 3.1.1.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario1.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario1.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.1.4 Simulate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100 # change here for number of replicates
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario1.rds")

```

#### 3.1.1.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario1.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.1.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.2 Scenario 2: 4 age-classes, same mortality across age-classes, mortality stationary over time and 0.5 noise to signal

#### 3.1.2.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # # logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.3),                                            # Mean Z on the log-scale
Z_CV = c(0.43),                                                 # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 3.1.2.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario2.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario2.RDS")

```

#### 3.1.2.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario2.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario2.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.2.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {
    
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }

}
# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario2.rds")

```

#### 3.1.2.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario2.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y =expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.2.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.3 Scenario 3: 4 age-classes, mortality decreases with age, mortality stationary over time and 0.5 noise to signal

#### 3.1.3.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                            # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(0.4002884, -0.3879585, -2.3710604, -2.3710604),   # Mean Z on the log-scale
Z_CV = c(0.124, 0.51, 1.02),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal 
,trend_type = "sinus"
)
```

#### 3.1.3.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario3.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario3.RDS")
```

#### 3.1.3.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario3.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario3.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.3.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {
    
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario3.rds")

```

#### 3.1.3.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario3.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.3.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.4 Scenario 4 - 4 age-classes, mortality increases with age, mortality stationary over time and 1 noise to signal

#### 3.1.4.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-2.3710604, -0.3879585,  0.4002884,  0.4002884),  # Mean Z on the log-scale
Z_CV = c(1.02, 0.51, 0.124),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 3.1.4.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario4.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario4.RDS")
```

#### 3.1.4.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario4.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario4.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.4.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {  
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}


# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario4.rds")

```

#### 3.1.4.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario4.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.4.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.5 Scenario 5: 4 age-classes, same mortality across age, mortality stationary over time and 1 noise to signal

#### 3.1.5.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # # logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.3),                                            # Mean Z on the log-scale
Z_CV = c(0.43),                                                 # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal   
,trend_type = "sinus"
)
```

#### 3.1.5.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario5.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario5.RDS")
```

#### 3.1.5.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario5.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario5.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.5.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {    
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}


# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario5.rds")

```

#### 3.1.5.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario5.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.5.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.6 Scenario 6: 4 age-classes, mortality decreases with age, mortality stationary over time and 1 noise to signal

#### 3.1.6.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                            # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(0.4002884, -0.3879585, -2.3710604, -2.3710604),   # Mean Z on the log-scale
Z_CV = c(0.124, 0.51, 1.02),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal
,trend_type = "sinus"
)
```

#### 3.1.6.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario6.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario6.RDS")
```

#### 3.1.6.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario6.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario6.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.6.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario6.rds")

```

#### 3.1.6.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario6.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.6.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.7 Scenario 7 - 7 age-classes, mortality increases with age, mortality stationary over time and 0.5 noise to signal

#### 3.1.7.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.7261684, -1.2659422, -1.5737547, -1.2047981, -0.5548759, -0.2927341, -0.2927341),  # Mean Z on the log-scale
Z_CV = c(2.07, 1.27, 1.09, 0.596, 0.319, 0.0283),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal    
,trend_type = "sinus"
)
```

#### 3.1.7.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario7.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario7.RDS")
```

#### 3.1.7.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario7.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario7.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.7.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario7.rds")

```

#### 3.1.7.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario7.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.7.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.8 Scenario 8 - 7 age-classes, same mortality across age, mortality stationary over time and 0.5 noise to signal

#### 3.1.8.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                         # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.24),                                            # Mean Z on the log-scale
Z_CV = c(0.61),   
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal
,trend_type = "sinus"
)
```

#### 3.1.8.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario8.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario8.RDS")
```

#### 3.1.8.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario8.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario8.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.8.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario8.rds")

```

#### 3.1.8.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario8.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.8.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.9 Scenario 9 - 7 age-classes, mortality decreases with age, mortality stationary over time and 1 noise to signal

#### 3.1.9.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.2927341, -0.5548759, -1.2047981, -1.5737547, -1.2659422, -1.7261684, -1.7261684),  # Mean Z on the log-scale
Z_CV = c(0.0283, 0.319, 0.596, 1.09, 1.27, 2.07),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 3.1.9.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario9.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario9.RDS")
```

#### 3.1.9.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario9.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario9.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.9.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario9.rds")

```

#### 3.1.9.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario9.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.9.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.10 Scenario 7 - 7 age-classes, mortality increases with age, mortality stationary over time and 1 noise to signal

#### 3.1.10.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.7261684, -1.2659422, -1.5737547, -1.2047981, -0.5548759, -0.2927341, -0.2927341),  # Mean Z on the log-scale
Z_CV = c(2.07, 1.27, 1.09, 0.596, 0.319, 0.0283),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 3.1.10.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario10.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario10.RDS")
```

#### 3.1.10.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario10.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario10.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.10.4 Simulate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario10.rds")

```

#### 3.1.10.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario10.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.10.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.11 Scenario 11 - 7 age-classes, same mortality across age, mortality stationary over time and 1 noise to signal

#### 3.1.11.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                         # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.24),                                            # Mean Z on the log-scale
Z_CV = c(0.61),   
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 3.1.11.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario11.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario11.RDS")
```

#### 3.1.11.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario11.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario11.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.11.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario11.rds")

```

#### 3.1.11.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario11.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.11.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.1.12 Scenario 12 - 7 age-classes, mortality decreases with age, mortality stationary over time and 1 noise to signal

#### 3.1.12.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.2927341, -0.5548759, -1.2047981, -1.5737547, -1.2659422, -1.7261684, -1.7261684),  # Mean Z on the log-scale
Z_CV = c(0.0283, 0.319, 0.596, 1.09, 1.27, 2.07),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 3.1.12.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario12.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario12.RDS")
```

#### 3.1.12.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario12.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario12.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.1.12.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario12.rds")

```

#### 3.1.12.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario12.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.1.12.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")
```

## 3.2 Increasing mortality

### 3.2.1 Scenario 13: 4 age-classes, mortality increases with age, mortality increases over time and 0.5 noise to signal

#### 3.2.1.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
logSD = c(0.8, 0.5, 0.2),                                       # SD on the log scale of mortality variability 
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-2.3710604, -0.3879585,  0.4002884,  0.4002884),  # Mean Z on the log-scale
Z_CV = c(1.02, 0.51, 0.124),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 3.2.1.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario13.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario13.RDS")
```

#### 3.2.1.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario13.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario13.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.1.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario13.rds")

```

#### 3.2.1.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario13.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.1.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.2 Scenario 14: 4 age-classes, same mortality across age-classes and 0.5 noise to signal

#### 3.2.2.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # # logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.3),                                            # Mean Z on the log-scale
Z_CV = c(0.43),                                                 # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 3.2.2.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario14.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario14.RDS")
```

#### 3.2.2.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario14.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario14.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.2.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {
    
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }

}
# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario14.rds")

```

#### 3.2.2.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario14.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.2.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.3 Scenario 15: 4 age-classes, mortality decreases with age, mortality increases over time and 0.5 noise to signal

#### 3.2.3.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                            # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(0.4002884, -0.3879585, -2.3710604, -2.3710604),   # Mean Z on the log-scale
Z_CV = c(0.124, 0.51, 1.02),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal 
,trend_type = "logistic"
)
```

#### 3.2.3.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario15.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario15.RDS")
```

#### 3.2.3.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario15.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario15.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.3.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {
    
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario15.rds")

```

#### 3.2.3.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario15.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.3.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.4 Scenario 16 - 4 age-classes, mortality increases with age, mortality increases over time and 1 noise to signal

#### 3.2.4.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-2.3710604, -0.3879585,  0.4002884,  0.4002884),  # Mean Z on the log-scale
Z_CV = c(1.02, 0.51, 0.124),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 3.2.4.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario16.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario16.RDS")
```

#### 3.2.4.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario16.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario16.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.4.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {  
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}


# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario16.rds")

```

#### 3.2.4.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario16.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.4.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.5 Scenario 17: 4 age-classes, same mortality across age, mortality increases over time and 1 noise to signal

#### 3.2.5.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # # logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.3),                                            # Mean Z on the log-scale
Z_CV = c(0.43),                                                 # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal   
,trend_type = "logistic"
)
```

#### 3.2.5.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario17.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario17.RDS")
```

#### 3.2.5.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario17.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario17.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.5.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {    
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}


# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario17.rds")

```

#### 3.2.5.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario17.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.5.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.6 Scenario 18: 4 age-classes, mortality decreases with age, mortality increases over time and 1 noise to signal

#### 3.2.6.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                            # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(0.4002884, -0.3879585, -2.3710604, -2.3710604),   # Mean Z on the log-scale
Z_CV = c(0.124, 0.51, 1.02),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal
,trend_type = "logistic"
)
```

#### 3.2.6.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario18.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario18.RDS")
```

#### 3.2.6.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario18.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario18.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.6.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario18.rds")

```

#### 3.2.6.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario18.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.6.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.7 Scenario 19 - 7 age-classes, mortality increases with age, mortality increases over time and 0.5 noise to signal

#### 3.2.7.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.7261684, -1.2659422, -1.5737547, -1.2047981, -0.5548759, -0.2927341, -0.2927341),  # Mean Z on the log-scale
Z_CV = c(2.07, 1.27, 1.09, 0.596, 0.319, 0.0283),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal    
,trend_type = "logistic"
)
```

#### 3.2.7.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario19.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario19.RDS")
```

#### 3.2.7.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario19.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario19.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.7.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario19.rds")

```

#### 3.2.7.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario19.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.7.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.8 Scenario 20 - 7 age-classes, same mortality across age, mortality increases over time and 0.5 noise to signal

#### 3.2.8.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                         # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.24),                                            # Mean Z on the log-scale
Z_CV = c(0.61),   
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal
,trend_type = "logistic"
)
```

#### 3.2.8.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario20.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario20.RDS")
```

#### 3.2.8.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario20.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario20.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.8.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario20.rds")

```

#### 3.2.8.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario20.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.8.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.9 Scenario 21 - 7 age-classes, mortality decreases with age, mortality increases over time and 1 noise to signal

#### 3.2.9.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.2927341, -0.5548759, -1.2047981, -1.5737547, -1.2659422, -1.7261684, -1.7261684),  # Mean Z on the log-scale
Z_CV = c(0.0283, 0.319, 0.596, 1.09, 1.27, 2.07),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 3.2.9.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario21.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario21.RDS")
```

#### 3.2.9.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario21.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario21.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.9.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario21.rds")

```

#### 3.2.9.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario21.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.9.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.10 Scenario 22 - 7 age-classes, mortality increases with age, mortality increases over time and 1 noise to signal

#### 3.2.10.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.7261684, -1.2659422, -1.5737547, -1.2047981, -0.5548759, -0.2927341, -0.2927341),  # Mean Z on the log-scale
Z_CV = c(2.07, 1.27, 1.09, 0.596, 0.319, 0.0283),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 3.2.10.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario22.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario22.RDS")
```

#### 3.2.10.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario22.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario22.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.10.4 Simulate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario22.rds")

```

#### 3.2.10.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario22.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.10.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.11 Scenario 23 - 7 age-classes, same mortality across age, mortality increases over time and 1 noise to signal

#### 3.2.11.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                         # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.24),                                            # Mean Z on the log-scale
Z_CV = c(0.61),   
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 3.2.11.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario23.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario23.RDS")
```

#### 3.2.11.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario23.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario23.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.11.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario23.rds")

```

#### 3.2.11.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario23.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.11.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 3.2.12 Scenario 24 - 7 age-classes, mortality decreases with age, mortality increases over time and 1 noise to signal

#### 3.2.12.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.2927341, -0.5548759, -1.2047981, -1.5737547, -1.2659422, -1.7261684, -1.7261684),  # Mean Z on the log-scale
Z_CV = c(0.0283, 0.319, 0.596, 1.09, 1.27, 2.07),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 3.2.12.2 Simulate and fit model to example data

```{r}
#| eval: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Save simulated data
saveRDS(sim,file = "../Data/Simulations/data_example_Scenario24.RDS")

# Save model fit
fit$save_object(file = "../Data/Simulations/fit_example_Scenario24.RDS")
```

#### 3.2.12.3 Comparison between simulated and estimated mortality rates

```{r}
# Load example data
sim <- readRDS("../Data/Simulations/data_example_Scenario24.rds")

# Load model fit to example data
fit <- readRDS("../Data/Simulations/fit_example_Scenario24.rds")

# compare simulated and estimated mortality rates
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 3.2.12.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: false

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario24.rds")

```

#### 3.2.12.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario24.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 3.2.12.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")
```

## 3.3 Summary of scenarios

### 3.3.1 Load and arrange all mortality simulations

```{r}
# load mortality data
file_paths <- list.files("../Data/Simulations", pattern="^Scenario", full.names = TRUE) # load all scenarios
file_paths <- file_paths[order(as.numeric(gsub("\\D", "", file_paths)))] # order file paths based on number inside strings

# load all scenarios
Z_all_scenarios <- lapply(file_paths,
                          readRDS)
names(Z_all_scenarios) <- file_paths

# bind all scenarios in one data frame
Z_all_scenarios <- bind_rows(Z_all_scenarios, .id="Scenario")

# Append info on all scenarios
Z_all_scenarios <-
Z_all_scenarios %>% 
  mutate(Mortality_time_pattern=ifelse(Scenario %in% c(1:12), "Stationary", "Increasing"),
         Number_of_age_classes=ifelse(Scenario %in% c(1:6,13:18),4,7),
         Mortality_age_relationship = case_when(Scenario %in% seq(1, 24, by=3) ~ "Increase",
                                                Scenario %in% seq(2, 24, by=3) ~ "Same",
                                                Scenario %in% seq(3, 24, by=3) ~ "Decrease",
                                                .default=NA
                                                ),
         Noise_vs_signal = ifelse(Scenario %in% c(1:3, 7:9, 13:15, 19:21), 0.5, 1)
         )

```

### 3.3.2 Make summaries

```{r}
Z_summary <-
Z_all_scenarios %>% 
  group_by(Z_type,Mortality_time_pattern, Number_of_age_classes, Mortality_age_relationship, Noise_vs_signal,Replicate) %>% 
  summarize(rmse=sqrt(mean(bias^2)),        # rmse for each rep multivariate Z ts.
            me=mean(bias),                  # mean error for each rep multivariate Z ts.
            re_rmse=sqrt(mean(rel_bias^2)), # rmse of rel bias for each rep mv Z ts.
            re_me = mean(rel_bias)          # mean rel error for each rep multivariate Z ts
            ) %>% 
  ungroup() %>% 
  group_by(Z_type,Mortality_time_pattern, Number_of_age_classes, Mortality_age_relationship, Noise_vs_signal) %>% 
  summarize(RMSE=mean(rmse),
            Bias=mean(me),
            sd_RMSE=sd(rmse),
            rel_RMSE = mean(re_rmse),
            rel_Bias = mean(re_me),
            sd_re_RMSE = sd(re_me) 
            ) %>%  # summarize over repl.
  ungroup() 


Z_summary_per_age_class <-
Z_all_scenarios %>% 
  group_by(Z_type,Mortality_time_pattern, Number_of_age_classes, Mortality_age_relationship, Noise_vs_signal, age_class, Replicate) %>% 
  summarize(rmse=sqrt(mean(bias^2)),        # rmse for each rep multivariate Z ts.
            me=mean(bias),                  # mean error for each rep multivariate Z ts.
            re_rmse=sqrt(mean(rel_bias^2)), # rmse of rel bias for each rep mv Z ts.
            re_me = mean(rel_bias)          # mean rel error for each rep multivariate Z ts
            ) %>% 
  ungroup() %>% 
  group_by(Z_type,Mortality_time_pattern, Number_of_age_classes, Mortality_age_relationship, Noise_vs_signal, age_class) %>% 
  summarize(RMSE=mean(rmse),
            Bias=mean(me),
            sd_RMSE=sd(rmse),
            rel_RMSE = mean(re_rmse),
            rel_Bias = mean(re_me),
            sd_re_RMSE = sd(re_me) 
            ) %>%  # summarize over repl.
  ungroup() 

```

### 3.3.3 Bias in time-varying mortality

```{r}
# Relative Bias and precision across scenarios for time_varying mortality
ggplot(Z_summary %>% filter(Z_type=="time_varying"), aes(x=rel_RMSE,y=sd_re_RMSE,colour= as.factor(Number_of_age_classes), shape= Mortality_age_relationship)) +
  geom_point() +
  ylab("Precision(sd(rmse))") +
  xlab("RMSE") +
  ggtitle("Relative bias") +
  facet_wrap(~Noise_vs_signal)

# classification tree relative bias
fit <- rpart(rel_RMSE ~ Mortality_time_pattern + Number_of_age_classes + Mortality_age_relationship + Noise_vs_signal, data= Z_summary_per_age_class %>% filter(Z_type=="time_varying"))
rpart.plot(fit, faclen=0)

# classification tree relative precision
fit <- rpart(sd_re_RMSE ~ Mortality_time_pattern + Number_of_age_classes + Mortality_age_relationship + Noise_vs_signal, data= Z_summary_per_age_class %>% filter(Z_type=="time_varying"))
rpart.plot(fit, faclen=0)




```

### 3. 3. 4 Bias in mean age-class specific mortality

```{r}
# Relative Bias and precision across scenarios for time_varying mortality
ggplot(Z_summary %>% filter(Z_type=="mean_mortality"), aes(x=rel_RMSE,y=sd_re_RMSE,colour= as.factor(Number_of_age_classes), shape= Mortality_age_relationship)) +
  geom_point() +
  ylab("Precision(sd(rmse))") +
  xlab("RMSE") +
  ggtitle("Relative bias") +
  facet_wrap(~Noise_vs_signal)

# classification tree relative bias
fit <- rpart(rel_RMSE ~ Mortality_time_pattern + Number_of_age_classes + Mortality_age_relationship + Noise_vs_signal, data= Z_summary_per_age_class %>% filter(Z_type=="mean_mortality"))
rpart.plot(fit, faclen=0)


pred <- predict(fit)
obs  <- fit$y
R2 <- 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
R2

# classification tree relative precision
fit <- rpart(sd_re_RMSE ~ Mortality_time_pattern + Number_of_age_classes + Mortality_age_relationship + Noise_vs_signal, data= Z_summary_per_age_class %>% filter(Z_type=="mean_mortality"))
rpart.plot(fit, faclen=0)


pred <- predict(fit)
obs  <- fit$y
R2 <- 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
R2

```

# 4. Conclusion

From simple time series plots it is clear that the model can not reproduce drastic fluctuations in mortality. Similar to random walks in general the model smooths fluctuations. It is further important to note that mortality is a latent parameter, that is, a variable that is not directly observed in the data but rather inferred from the model using abundance data. Hence, we are not to expect a "perfect" fit to data. Notwithstanding, the model often gives plausible estimates of simulated mortality rates and mimics temporal patterns. It is clear, both when comparing temporal patterns in observed and estimated age-class specific mortality rates, when comparing estimated and observed values, and when comparing bias to observed values, that the model tends to underestimate high mortality values and overestimate low mortality rates.

Mean age-class specific mortality rates are well reproduced by the model, something which is clearly visible in plots showing bias in mean age-class specific mortality and the relatively good coverage that credibility intervals of mean age-class mortality have on observed values. Mean age-class specific mortality rates are better reproduced in the case of 7 age-classes as compared to 4 age-classes.

Below follows some specific comments, comparing the different scenarios that were tested:

1.  Stationary mortality dynamics vs. trend

-   It is unclear to what extent the model handles stationary or trends in a better way. From visual inspections of bias it seems that the model is better at reproducing trends. However, most important the model can clearly handle trends in mortality, something which is observed in the model systems.\

2.  Number of age-classes

-   It appears as if bias is lower in the case of 7 vs. 4 age classes.

3.  Mortality increase, the same or decrease with age

-   It is clear that model can handle all of these cases. If the observed mortality increase with age so does mortality estimates. **HENCE**, the increase in mean mortality with age observed in the example systems is not a model artifact.

4.  Noise to signal ratio

-   An increasing noise to signal ratio in observed mortality naturally leads to a large bias in parameter estimates. This is expected, the random walk just follows the mean pattern.

*Overall, this simulation experiment clearly shows that model estimates are plausible!!*

-   Stationary vs. deterministic trend
