---
title: "Simulation_testing_of_perch_model"
author: "Torbjörn Säterberg"
date: today  
format: html
embed-resources: true
self-contained: true
editor: visual
code-fold: true
error: FALSE
warning: FALSE
message: FALSE
toc: true
toc-depth: 3
lang: en  
---

# 1. Introduction

This script is used for testing the reliability of age-based perch model using data simulations. With this regards two general analyses are conducted:

-   First, we estimate bias in age-based mortality parameters using simulated data.

-   Second, as there are indications that large perch tend to uphold in shallower waters than what smaller perch do we subset the data based on water depth at which gillnets are set and redo the analysis using various data subsets.

```{r}
library(MASS)  # Stats functions
library(tidyverse) # tidy functions
library(cmdstanr) 
```

# 2. Simulation based bias estimation

```{r}

# List all .R files in the directory
files.sources <- list.files(path = "../R-simulations", pattern = "\\.R$", full.names = TRUE)

# Source each file
sapply(files.sources, source)
```

## 2.1 Stationary mortality

Scenarios

| Scenario | Number of age-classes | Mortality pattern        | Noise vs signal |
|----------|-----------------------|--------------------------|-----------------|
| S1       | 4                     | Increase with age        | 0.5             |
| S2       | 4                     | Same for all age-classes | 0.5             |
| S3       | 4                     | Decrease with age        | 0.5             |
| S4       | 4                     | Increase with age        | 1               |
| S5       | 4                     | Same for all age-classes | 1               |
| S6       | 4                     | Decrease with age        | 1               |
| S7       | 7                     | Increase with age        | 0.5             |
| S8       | 7                     | Same for all age-classes | 0.5             |
| S9       | 7                     | Decrease with age        | 0.5             |
| S10      | 7                     | Increase with age        | 1               |
| S11      | 7                     | Same for all age-classes | 1               |
| S12      | 7                     | Decrease with age        | 1               |

: Simulation scenarios {#tbl-sim}

### 2.1.1 Scenario 1: 4 age-classes, mortality increases with age, mortality stationary over time and 0.5 noise to signal

#### 2.1.1.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
logSD = c(0.8, 0.5, 0.2),                                       # SD on the log scale of mortality variability 
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-2.3710604, -0.3879585,  0.4002884,  0.4002884),  # Mean Z on the log-scale
Z_CV = c(1.02, 0.51, 0.124),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 2.1.1.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.1.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.1.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario1.rds")

```

#### 2.1.1.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario1.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.1.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.2 Scenario 2: 4 age-classes, same mortality across age-classes, mortality stationary over time and 0.5 noise to signal

#### 2.1.2.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # # logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.3),                                            # Mean Z on the log-scale
Z_CV = c(0.43),                                                 # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 2.1.2.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.2.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.2.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {
    
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }

}
# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario2.rds")

```

#### 2.1.2.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario2.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.2.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.3 Scenario 3: 4 age-classes, mortality decreases with age, mortality stationary over time and 0.5 noise to signal

#### 2.1.3.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                            # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(0.4002884, -0.3879585, -2.3710604, -2.3710604),   # Mean Z on the log-scale
Z_CV = c(0.124, 0.51, 1.02),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal 
,trend_type = "sinus"
)
```

#### 2.1.3.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.3.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.3.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {
    
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario3.rds")

```

#### 2.1.3.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario3.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.3.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.4 Scenario 4 - 4 age-classes, mortality increases with age, mortality stationary over time and 1 noise to signal

#### 2.1.4.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-2.3710604, -0.3879585,  0.4002884,  0.4002884),  # Mean Z on the log-scale
Z_CV = c(1.02, 0.51, 0.124),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 2.1.4.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.4.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.4.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {  
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}


# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario4.rds")

```

#### 2.1.4.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario4.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.4.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.5 Scenario 5: 4 age-classes, same mortality across age, mortality stationary over time and 1 noise to signal

#### 2.1.5.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # # logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.3),                                            # Mean Z on the log-scale
Z_CV = c(0.43),                                                 # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal   
,trend_type = "sinus"
)
```

#### 2.1.5.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.5.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.5.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {    
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}


# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario5.rds")

```

#### 2.1.5.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario5.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.5.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.6 Scenario 6: 4 age-classes, mortality decreases with age, mortality stationary over time and 1 noise to signal

#### 2.1.6.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                            # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(0.4002884, -0.3879585, -2.3710604, -2.3710604),   # Mean Z on the log-scale
Z_CV = c(0.124, 0.51, 1.02),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal
,trend_type = "sinus"
)
```

#### 2.1.6.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.6.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.6.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario6.rds")

```

#### 2.1.6.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario6.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.6.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.7 Scenario 7 - 7 age-classes, mortality increases with age, mortality stationary over time and 0.5 noise to signal

#### 2.1.7.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.7261684, -1.2659422, -1.5737547, -1.2047981, -0.5548759, -0.2927341, -0.2927341),  # Mean Z on the log-scale
Z_CV = c(2.07, 1.27, 1.09, 0.596, 0.319, 0.0283),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal    
,trend_type = "sinus"
)
```

#### 2.1.7.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.7.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.7.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario7.rds")

```

#### 2.1.7.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario7.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.7.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.8 Scenario 8 - 7 age-classes, same mortality across age, mortality stationary over time and 0.5 noise to signal

#### 2.1.8.1 Set simulation parameters
```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                         # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.24),                                            # Mean Z on the log-scale
Z_CV = c(0.61),   
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal
,trend_type = "sinus"
)
```

#### 2.1.8.2 Simulate and fit model to example data
```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.8.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.8.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario8.rds")

```

#### 2.1.8.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario8.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.8.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.9 Scenario 9 - 7 age-classes, mortality decreases with age, mortality stationary over time and 1 noise to signal


#### 2.1.9.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.2927341, -0.5548759, -1.2047981, -1.5737547, -1.2659422, -1.7261684, -1.7261684),  # Mean Z on the log-scale
Z_CV = c(0.0283, 0.319, 0.596, 1.09, 1.27, 2.07),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 2.1.9.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.9.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.9.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario9.rds")

```

#### 2.1.9.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario9.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.9.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.10 Scenario 7 - 7 age-classes, mortality increases with age, mortality stationary over time and 1 noise to signal

#### 2.1.10.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.7261684, -1.2659422, -1.5737547, -1.2047981, -0.5548759, -0.2927341, -0.2927341),  # Mean Z on the log-scale
Z_CV = c(2.07, 1.27, 1.09, 0.596, 0.319, 0.0283),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 2.1.10.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.10.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.10.4 Simulate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario10.rds")

```

#### 2.1.10.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario10.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.10.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.11 Scenario 11 - 7 age-classes, same mortality across age, mortality stationary over time and 1 noise to signal

#### 2.1.11.1 Set simulation parameters
```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                         # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.24),                                            # Mean Z on the log-scale
Z_CV = c(0.61),   
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 2.1.11.2 Simulate and fit model to example data
```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.11.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.11.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario11.rds")

```

#### 2.1.11.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario11.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.11.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.1.12 Scenario 12 - 7 age-classes, mortality decreases with age, mortality stationary over time and 1 noise to signal

#### 2.1.12.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.2927341, -0.5548759, -1.2047981, -1.5737547, -1.2659422, -1.7261684, -1.7261684),  # Mean Z on the log-scale
Z_CV = c(0.0283, 0.319, 0.596, 1.09, 1.27, 2.07),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "sinus"
)
```

#### 2.1.12.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.1.12.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.1.12.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario12.rds")

```

#### 2.1.12.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario12.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.1.12.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")
```

## 2.2 Increasing mortality

### 2.2.1 Scenario 13: 4 age-classes, mortality increases with age, mortality increases over time and 0.5 noise to signal

#### 2.2.1.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
logSD = c(0.8, 0.5, 0.2),                                       # SD on the log scale of mortality variability 
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-2.3710604, -0.3879585,  0.4002884,  0.4002884),  # Mean Z on the log-scale
Z_CV = c(1.02, 0.51, 0.124),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 2.2.1.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.1.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.1.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario13.rds")

```

#### 2.2.1.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario13.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.1.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.2 Scenario 14: 4 age-classes, same mortality across age-classes and 0.5 noise to signal

#### 2.2.2.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # # logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.3),                                            # Mean Z on the log-scale
Z_CV = c(0.43),                                                 # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 2.2.2.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.2.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.2.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {
    
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }

}
# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario14.rds")

```

#### 2.2.2.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario14.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.2.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.3 Scenario 15: 4 age-classes, mortality decreases with age, mortality increases over time and 0.5 noise to signal

#### 2.2.3.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                            # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(0.4002884, -0.3879585, -2.3710604, -2.3710604),   # Mean Z on the log-scale
Z_CV = c(0.124, 0.51, 1.02),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal 
,trend_type = "logistic"
)
```

#### 2.2.3.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.3.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.3.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {
    
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario15.rds")

```

#### 2.2.3.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario15.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.3.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.4 Scenario 16 - 4 age-classes, mortality increases with age, mortality increases over time and 1 noise to signal

#### 2.2.4.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-2.3710604, -0.3879585,  0.4002884,  0.4002884),  # Mean Z on the log-scale
Z_CV = c(1.02, 0.51, 0.124),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 2.2.4.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.4.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.4.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {  
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}


# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario16.rds")

```

#### 2.2.4.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario16.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.4.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.5 Scenario 17: 4 age-classes, same mortality across age, mortality increases over time and 1 noise to signal

#### 2.2.5.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # # logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.3),                                            # Mean Z on the log-scale
Z_CV = c(0.43),                                                 # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal   
,trend_type = "logistic"
)
```

#### 2.2.5.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.5.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.5.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){


success <- FALSE

while (!success) {    
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}


# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario17.rds")

```

#### 2.2.5.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario17.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.5.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.6 Scenario 18: 4 age-classes, mortality decreases with age, mortality increases over time and 1 noise to signal

#### 2.2.6.1 Set simulation parameters

Here we define parameters based on scenario 1 in @tbl-sim.

```{r}
sim_params <- list(
mu_log_recruitment = 0.662,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 1.22,                                      # SD of recruitment on log-scale
n_age_classes = 4,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(3.2800550, 1.2601100, 0.6301085, 1.0629100),         # Abundance at initial time step for each age-class
NBphi = 1.388495,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                            # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(0.4002884, -0.3879585, -2.3710604, -2.3710604),   # Mean Z on the log-scale
Z_CV = c(0.124, 0.51, 1.02),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal
,trend_type = "logistic"
)
```

#### 2.2.6.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.6.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.6.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario18.rds")

```

#### 2.2.6.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario18.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.6.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.7 Scenario 19 - 7 age-classes, mortality increases with age, mortality increases over time and 0.5 noise to signal

#### 2.2.7.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.7261684, -1.2659422, -1.5737547, -1.2047981, -0.5548759, -0.2927341, -0.2927341),  # Mean Z on the log-scale
Z_CV = c(2.07, 1.27, 1.09, 0.596, 0.319, 0.0283),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal    
,trend_type = "logistic"
)
```

#### 2.2.7.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.7.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.7.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario19.rds")

```

#### 2.2.7.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario19.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.7.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.8 Scenario 20 - 7 age-classes, same mortality across age, mortality increases over time and 0.5 noise to signal

#### 2.2.8.1 Set simulation parameters
```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                         # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.24),                                            # Mean Z on the log-scale
Z_CV = c(0.61),   
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal
,trend_type = "logistic"
)
```

#### 2.2.8.2 Simulate and fit model to example data
```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.8.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.8.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario20.rds")

```

#### 2.2.8.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario20.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.8.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.9 Scenario 21 - 7 age-classes, mortality decreases with age, mortality increases over time and 1 noise to signal


#### 2.2.9.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.2927341, -0.5548759, -1.2047981, -1.5737547, -1.2659422, -1.7261684, -1.7261684),  # Mean Z on the log-scale
Z_CV = c(0.0283, 0.319, 0.596, 1.09, 1.27, 2.07),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 0.5                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 2.2.9.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.9.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.9.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario21.rds")

```

#### 2.2.9.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario21.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.9.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.10 Scenario 22 - 7 age-classes, mortality increases with age, mortality increases over time and 1 noise to signal

#### 2.2.10.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.7261684, -1.2659422, -1.5737547, -1.2047981, -0.5548759, -0.2927341, -0.2927341),  # Mean Z on the log-scale
Z_CV = c(2.07, 1.27, 1.09, 0.596, 0.319, 0.0283),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 2.2.10.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.10.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.10.4 Simulate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario22.rds")

```

#### 2.2.10.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario22.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.10.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.11 Scenario 23 - 7 age-classes, same mortality across age, mortality increases over time and 1 noise to signal

#### 2.2.11.1 Set simulation parameters
```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                         # Number of gillnet-nights per year
age_class_specific_mortality = FALSE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-1.24),                                            # Mean Z on the log-scale
Z_CV = c(0.61),   
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 2.2.11.2 Simulate and fit model to example data
```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.11.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.11.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario23.rds")

```

#### 2.2.11.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario23.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.11.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")

```

### 2.2.12 Scenario 24 - 7 age-classes, mortality decreases with age, mortality increases over time and 1 noise to signal

#### 2.2.12.1 Set simulation parameters

```{r}
sim_params <- list(
mu_log_recruitment = 1.32,                                     # Expected value of recruitment on log-scale
sd_log_recruitment = 0.60,                                      # SD of recruitment on log-scale
n_age_classes = 7,                                              # Number of age-classes in the model
t_max = 20,                                                     # Number of time steps to simulate
N_init = c(4.70189500, 7.76934500, 9.50462500, 1.42273000, 0.19977400, 0.09031005, 0.20990450),         # Abundance at initial time step for each age-class
NBphi = 2.69573,                                               # Dispersion parameter of negative binomial
n_gillnet_nights = 30,                                          # Number of gillnet-nights per year
age_class_specific_mortality = TRUE,                           # Logical whether different mortality rates per age-class(TRUE) or mortality assumed to be the same for all age-classes(FALSE)
logZ_mean = c(-0.2927341, -0.5548759, -1.2047981, -1.5737547, -1.2659422, -1.7261684, -1.7261684),  # Mean Z on the log-scale
Z_CV = c(0.0283, 0.319, 0.596, 1.09, 1.27, 2.07),                                    # CV of Z on ordinary scale
noise_vs_sinus_signal = 1                                     # How large should noise (normal)be in comparison to sinusoidal signal     
,trend_type = "logistic"
)
```

#### 2.2.12.2 Simulate and fit model to example data

```{r}
#| include: false

# simulate data
sim <- simulate_perch_model(sim_params)              

# compile model
model <- cmdstan_model('../STAN/age_model_logN_Z_rw_STAN.stan')

# fit model
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)
```

#### 2.2.12.3 Comparison between simulated and estimated mortality rates

```{r}
compare_simulated_and_estimated_mortality_rates(fit, sim$Z)
```

#### 2.2.12.4 Simualate data for bias assessment (Obs slow!!!)

```{r}
#| eval: true

n_rep <- 100
Z <- tibble()

# loop over all replicates
for(i in c(1:n_rep)){

success <- FALSE

while (!success) {      
  
# Simulate data
sim <- simulate_perch_model(sim_params)

# Fit model to simulated data
fit <- model$sample(
  data = sim$STAN_input,
  iter_warmup = 1000, 
  iter_sampling = 1000,
  init = generate_inits_list(sim$STAN_input,chains=4),
  adapt_delta=0.8,
  seed = 125,
  chains = 4,
  parallel_chains = 4,
  refresh = 100
)

    # check if chains ran
    if (length(fit$output_files()) != 0) {
      success <- TRUE
    } else {
      message(paste("Retrying model", i))
    }
}

# Merge estimated(median of the posterior distribution) and simulated data
Z_repl <- extract_Z_estimates_and_simulated_true_parameters(fit, sim$Z) %>%
          add_column(Replicate=i,.before="Z_est") # add replicate number

# Append results from single replicate to larger data.frame
Z <- bind_rows(Z, Z_repl)

print(i)
}

# save data for later use
saveRDS(Z, file="../Data/Simulations/Scenario24.rds")

```

#### 2.2.12.5 Bias in timevarying mortality

```{r}
# load mortality data
Z <- readRDS("../Data/Simulations/Scenario24.rds")

################################################################################
#------------------------- Time varying mortality -----------------------------#
################################################################################
Z_time <- Z %>% filter(Z_type=="time_varying")

# True vs. Estimated values
ggplot(Z_time, aes(x = Z_true, y = Z_est)) + geom_point() + geom_abline(intercept=0, slope=1) + labs(x=expression("Observed " (theta)), y=expression("Estimated " (hat(theta)))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Violin plot of bias
ggplot(Z_time, aes(x=age_class, y = bias)) + geom_violin() + labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) + theme_bw() 

# Histogram of bias
ggplot(Z_time, aes(x=bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + 
  labs(x = expression("Bias " (hat(theta) - theta))) + facet_wrap(~age_class) + theme_bw()

# Violin plot relative bias
ggplot(Z_time, aes(x=age_class, y=rel_bias)) + geom_violin() + labs(y = expression("Relative bias " (hat(theta) - theta) / theta), x = "Age class") + theme_bw()

# Histogram of relative bias
ggplot(Z_time, aes(x=rel_bias)) + geom_histogram() + geom_vline(xintercept = 0, colour="red") + labs(x = expression("Relative bias " (hat(theta) - theta) / theta)) + facet_wrap(~age_class) + theme_bw()

# Observed vs bias
ggplot(Z_time, aes(x = Z_true, y = bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

# Observed vs relative bias
ggplot(Z_time, aes(x = Z_true, y = rel_bias)) + geom_point() + geom_hline(yintercept = 0, colour="red") + labs(x=expression("Observed " (theta)), y = expression("Bias " (hat(theta) - theta))) + facet_wrap( ~ age_class, scales="free") + theme_bw()

```

#### 2.2.12.6 Bias in mean age-class mortality

```{r}
################################################################################
#------------------------------ Mean mortality --------------------------------#
################################################################################
Z_mean <- Z %>% filter(Z_type=="mean_mortality")

Z_mean %>% 
  pivot_longer(cols=c(Z_est,Z_true), names_to = "obs_or_est", values_to = "Z") %>%
ggplot(., aes(y=Z, x=obs_or_est, fill=obs_or_est)) + 
  labs(x = "Observed and estimated Z", y = "Z") +
  geom_violin() +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.5,
    linewidth = 0.4
  ) +
  theme_bw() + 
  facet_wrap(~age_class, scales="free")

ggplot(Z_mean, aes(y=bias, x=age_class)) + 
  labs(x = "Age class", y = expression("Bias " (hat(theta) - theta))) +
  geom_boxplot() +
  geom_hline(yintercept=0, colour="red") +
  theme_bw()

ggplot(Z_mean, aes(x = age_class, fill = coverage)) +
  geom_bar(position="fill") +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No"),
    name = "CI covers true value"
  ) +
  labs(x="Age class", 
       y="Proportion cases CI includes or not includes true value")
```


# 3. Effect of data subsetting on
